---
title: "R basics part 1"
subtitle: "Explain script sections , the here package, variable assignments and functions"
author: "Han Olff"
date: "2026-8-29"
editor_options: 
  chunk_output_type: console
execute:
  output: false
  warning: false
  message: false
---

# Quarto scripts and R scripts

Quarto is a publishing system that allows you to create documents, reports, and presentations in various formats. R scripts are used to perform data analysis and visualization in R. Quarto scripts can include R code chunks, which are executed when the document is rendered. Quarto scripts are also usefull for explaining to yourself or to others (as we will do in this course) how your R scripts work. Quarto scripts can be rendered to HTML, PDF, or Word documents, while R scripts are typically run in an interactive R session. Quarto scripts can also include text, images, and other media, while R scripts are primarily focused on code and output.
Both quarto and R scripts are plain text files, and can be edited in any text editor. However, RStudio provides a user-friendly interface for working with both types of scripts, including syntax highlighting, code completion, and debugging tools. When the extension of a script is .r it is treated as an R script, and when the extension is .qmd it is treated as a Quarto script. In this course we will use Quarto scripts to explain how R scripts work, and to provide additional context and information about the analysis.

# 02 Collapsing or expanding sections of your script, and using an outline of your script

In a quarto script, you can use headers to create sections. Use `#` for a top-level header, `##` for a second-level header, and so on. The number of `#` symbols indicates the level of the header. For example, `# 01 Setup your environment` is a top-level header, while `## 01.1 Install packages` is a second-level header. The headers can be collapsed or expanded in the script editor, allowing you to hide or show the content of each section. This is useful for organizing your code and making it easier to navigate. Collapse a sections using the triangle next to the header, and expand it again by clicking on the triangle again. You can also use the keyboard shortcut `Ctrl + Shift + -` to collapse all sections, and `Ctrl + Shift + =` to expand all sections.
When a line in an R script starts with `----` (4 or more `-`), you can also collapse and expand it. Combine this with numbering the sections of your script, and always make section 01 the header. 

Headers in R or Quarto scripts allow you to show an outline of your script — in RStudio use Code/Show Document Outline — and allows quick navigation through your script. In the Explorer panel on the left in Positron, you can view the outline of your script that uses the headers, while also showing defined variables and functions in each section of your script.
+
# 03 Using renv for making R predictable

A major strength, but also a potential challenge, of R is that the software changes all the time. All functionality of R comes from packages, and these are improved, expanded, and updated all the time. Say that you develop a set of scripts in R now, for example to analyse the data in your master project, and you develop a publication from the results. But you get several revisions, and finally complete the paper two years later. By that time, you will have updated your versions of R and your packages several times. This may change the functionality of your scripts, so they may not work anymore. And when you publish the paper and deposit the data and scripts in a repository, even 10 years from now you want to be able to replicate your analyses, with the "old" versions of the packages that you used at the time when you did the analysis for the paper.

A similar problem arises when working on scripts and data in a collaborative project with multiple people. Typically, not every collaborator will have exactly the same packages and same versions of the same package installed in their R library. This means only one person can improve the scripts and run them, instead of everyone contributing.

The renv package solves this in combination with using an RStudio project and Git/GitHub — for the basis see https://rstudio.github.io/renv/. In a "fresh" project, you first create an `renv.lock` file with `renv::init()`. You typically only do this once. This initialization of your project serves two goals:

1.  new packages are installed from then on in a library in the local folder of your project, not anymore in your general R installation, so the packages "belong" to the project.
2.  you, or your collaborators, can use `renv::restore()` to create now or later exactly the same R library (same versions of the packages) as you are using to develop the scripts, where these version numbers are read from the `renv.lock` file. If certain packages are not installed yet on that specific computer, they will be installed.

This ensures that your project uses the exact same package versions and dependencies that were initially recorded, making it easier to reproduce analyses and maintain consistency across different environments or collaborators. The version of packages in the library can be updated by yourself or your collaborator, but then `renv::restore()` causes everyone to use the same versions of all libraries. Be careful with this: only update your library when really necessary (e.g. a package contains an error or missing functionality). Otherwise just stick to the version that you used when you started the project. You do not need to do `renv::init()` now, because you have cloned an existing project.

Make sure that before you use `renv::restore()` on a Windows machine, you have RTools installed. This is a set of tools that are needed to compile some packages from source. Install the RTools version that matches your R version — so if you have R version 4.6.x you need RTools45 (it works with R 4.5 and R 4.6). You can download RTools from the CRAN website at https://cran.r-project.org/bin/windows/Rtools, and choose the Rtools installer, using default installation settings. Note that this works differently on a Mac: there you do not work with RTools, but with Apple's developer toolchain instead, see https://mac.r-project.org/tools/.

# 04 Setup your working environment

Run the setup script, containing package restoration, helper functions, etc.

```{r}
# run the setup script
source(here::here("scripts", "01-setup.R"))

# pull the changes in the online repository to this local folder
# this is important when you collaborate on the project
system2("git", c("-C", here::here(), "pull"))

# show which package namespaces are loaded to memory with library()
search()
```

# 05 The here package

The `here` package gets the root folder of your project; use this instead of absolute file locations. For example, do NOT use `source("c:\github\ERS2026\01-setup.R")`, as this will cause problems when collaborating on your project.

```{r}
# show the root folder of your git-enabled project
here::here()
```

# 06 Assignments, calculations, functions, vectors, lists, data frames

## Assigning objects, and using functions from a core package

Objects are anything stored in R in memory that you can refer to by a name. Create objects through the assignment operator `<-` (variables). Functions in R perform calculations and are typically part of a package. R comes with a set of core packages installed with the default R installation — mainly `base`, `stats`, `utils`, `graphics`, `grDevices`, `methods`, and `datasets`. Functions in these core packages can be used without loading the package with `library()`; for example, `sum()` is in the `base` package.

```{r}
# adds two numbers 
1 + 3

# assign two variables and add them
var1 <- 1
var2 <- 3
var1 + var2

# use built-in function from a package
sum(var1, var2)

# Objects can hold more than one value. Vectors are data structures with more than one value, create them with the concatenate (c()) function
days <- c(1, 2, 3, 4, 5, 6)
days
class(days)
bodymass_g <- c(2, 3, 5, 7, 9, 11)
base::plot(days, bodymass_g)

# Note that the default for missing values is na.rm = FALSE, so if one of the values is NA, the result will be NA. You can change this by setting na.rm = TRUE.
var3<-NA
data<-c(var1, var2, var3)
data
sum(data)
sum(data, na.rm = TRUE)

# a full specification for using the sum function would be
base::sum(data, na.rm = TRUE)
# but for core packages such as base (that come with the default R
# installation) you typically do not specify from which package the
# function comes
```

## Inspecting the code for a built-in function (from a core package)

```{r}
a <- c(1, 2, 3)
b <- 10
base::append(a, b)
base::append
```

```{r}
#| eval: false
?base::append()
```

## Data types and data structures in R

R has different data types, such as numeric, character, logical, factor and list. You can check the data type of an object with the `class()` function. Data can be organized in different data structures, such as vectors, matrices, data frames and lists. You can check the data structure of an object with the `str()` function.

## Named lists

A named list is a collection of different objects, which can be of different types and lengths, such as text, numbers, or boolean variables. Create a named list with four elements.

```{r}
elephant_properties <- list(
  name = "Elephant",
  weight = 5400,
  is_endangered = TRUE,
  colors = c("gray", "white")
)
class(elephant_properties)
# show the elements of the list
names(elephant_properties)
# inspect the contents of the elements
elephant_properties

# show only the weight
elephant_properties$weight
```

A data frame is a special type of named list with rows and columns (variables). In practice, we typically read a data frame from a file, rather than creating it ourselves in a script like this.

```{r}
students <- data.frame(
  name = c("Alex", "Sam", "Robin"),
  age = c(21, 23, 20)
)
print(students)
barplot(age ~ name, data = students)
```

# 07 Using non-core package functions

For functions from non-core packages, always specify `package::function(parameter1, parameter2)`, because the same function may exist in different packages.

```{r}
#| eval: false
# show help on the lubridate package
help(package = lubridate)
```

```{r}
# list the functions in the lubridate package
ls("package:lubridate")
```

```{r}
#| eval: false
# get help on the function floor_date from lubridate
help("floor_date", package = "lubridate")
```

```{r}
# calculate the date of the monday of the same week of a date
lubridate::floor_date(lubridate::dmy("29-Aug-2026"),
  unit = "week",
  week_start = 1
)
```

# 08 Writing your own functions

Your new function can use existing functions from packages — here the function `dmy()` is used from the lubridate package.

```{r}
# define the function days_between: the number of days between two dates
days_between <- function(from, to) {
  as.integer(lubridate::dmy(to) - lubridate::dmy(from))
}
# use the function for a calculation
days_between("1-Apr-2026", "31-Aug-2026")
```

```{r}
# Define a function that calculates the area of a circle.
circle_area <- function(radius) { # radius is an argument of the function
  area <- pi * radius^2 # calculate area as A = pi * r^2
  return(area) # return the calculated value
}
# Call the function: calculate the area of a circle with radius = 5
circle_area(5) # the function returns the area
```

# 09 Datasets in packages

Most packages contain especially functions, but sometimes also (example) datasets.

```{r}
# list what the package dplyr contains
ls("package:dplyr")

# print the dataset starwars dataframe that is in the dplyr package
class(dplyr::starwars)
print(dplyr::starwars)

# show all variable names of the dataset
names(dplyr::starwars)

# show all levels of the variable name as a vector
dplyr::starwars$name

# frequency table of home world
table(dplyr::starwars$homeworld)
```
