---
title: "03-organizing and importing data"
author: "Han Olff"
date: "2025-8-22"
format: html
editor: 
  markdown: 
    wrap: 72
editor_options: 
  chunk_output_type: console
---

How to import data into R from different sources, including local files and online databases.
You will learn to not use local data files, but instead read the data from an online Google Sheets database

# 01 Setup your environment

Run the common setup script and synchronize the local project with
GitHub.

```{r}
# Run the setup script containing package restoration, helper functions, etc.
source(here::here("scripts", "01-setup.R"))

# Pull changes from the online repository into this local project folder.
# This is especially important when collaborating on the same project.
system2("git",c("-C", here::here(), "pull"))

# show which package namespaces are loaded to memory with library()
search()

# authenticate your @student.rug.nl google account
gsheets_auth()
```

# 02 Different ways to enter data

Data can be imported into R in different ways:

Method 1 — entering data directly in your script
Method 2 — reading data from a local `.csv` or `.xlsx` file on your computer
Method 3 — reading data from a published csv link to a Google Sheet

We will mostly use the last method during this course. When reading
data, we prefer the readr::read_csv() function, as it reads data
directly into a tibble instead of a dataframe. A tibble is more
compact when printed and shows the variable types directly.

# 03 Method 1 - enter data directly in your script

This is only practical for small datasets, but not for large datasets.
We did this already in the previous scripts.

# 04 Method 2 - read data from a local file

If you want to read a file from a drive from your computer, it is a
good idea to set a working directory to point where your different
datasets are located.

```{r}
EVI2001_2023 <- readr::read_csv(
  here::here("data", "EVI2001_2023.csv")
)
print(EVI2001_2023)
```

This reads the file EVI2001_2023.csv from the folder data on my local
folder. Although it is synchronized through git/github, that is only
possible for small files, and also it lacks all kind of metadata, it
is a loose file, not a database. Also it is not a good idea to keep
data outside the git folder, as then we are not sure we work on the
same data.

# 05 Method 3 - read data from  online databases

This is the highly recommended choice of these three methods.
This means that there is only one version of
the data, and all collaborators read the same data. This is therefore
preferred method for collaborative projects, especially when the data change
 over time. The all team members always work with the up-to-date data.

Different online database options exist, ranging from a MySQL or
PostgreSQL database on a server to a spreadsheet-based cloud
database such as Google Sheets. This course uses Google Sheets
because it is easy for a team to set up and access. It is flexible,
but that flexibility is also a drawback — mistakes are easy to make,
so it is important to structure and document the database clearly,
as explained in the next sections.

In research projects beyond this course, a dedicated
PostgreSQL/PostGIS database on a server is often preferable: it
enforces data quality more strictly and can run complex queries on
the server rather than in the R script.

Also for projects where you are the only one to work with the data,
this is the preferred methods. It avoids the problem of having
multiple versions of the same data on different computers. It also
allows you to read the data directly into R, without the need to
download and store a local copy of the data. This is especially
important for large datasets, as it avoids the need to store large
files on your computer. It also allows you to read the data directly
into R, without the need to download and store a local copy of the
data. 

Google Sheets is a spreadsheet app, similar to Microsoft Excel, but its
files "live" in the cloud, associated with your Google account.
Although a spreadsheet app is formally not a database app, a Google
Sheets document can be set up as a lightweight relational database
when organized according to a few clear principles (as described
here).

A key feature is that you can read data in Google Sheets directly into R, 
similar to using a database server.
To do this, use the function read_gsdb() that you defined in the
script 01-setup.r. This function reads in all tables (sheets) in a
Google Sheets database and returns a named list object, where the
elements of the object are different tibbles. You can then access
each tibble using the $ operator, e.g.
transdat$FactSectionAnimals.
Authentication makes sure that you can set that only specific people
can read the data, and that you can set who can edit the data. This
is important for collaborative projects, where you want to allow
some people to edit the data, but not others. It is also important
to set the correct permissions for each table (sheet) in the
database, as some tables may contain sensitive data that should not
be shared with everyone.

Reading data this way keeps all data management in one  (online) place: it's
best practice to give the R dataframe the same name as the online
sheet, so the same script can be run on any computer and all
collaborators read identical data.

# 06 Introduction example online database

We explore this with an example Google Sheets database. 
Open in it in your browserwith crtl + click on the next link:
https://docs.google.com/spreadsheets/d/1m-liu8omZMewqz_YP9j_YUmQ0zwATl3z4aRLnZFnfWc/edit?usp=sharing

This is a Google Sheets database that only exists in one location in
the cloud, so all of us view the same database - there is only one
version of the data. It is a relational database, with multiple
tables (sheets) that are linked to each other. The database is set up
as a Star Schema, which is a particular way of organising data in a
relational database. The database contains data on a study on the
effects of wildlife on vegetation in East Africa. The database
contains the following tables (sheets):

-   MetStudyInfo: short descriptions of the key features of your study
-   MetStill2Do: list of what you still need to do, your ToDo list for
    the database
-   MetTables: the list of tables (sheets) in your database, with for
    each table a short contents description
-   DimTransect: the list of transects in the study, with their
    properties
-   DimPlot: the list of plots in the study, with their properties
-   DimSpecies: the list of species in the study, with their
    properties
-   DimSection: the list of sections in the study, with their
    properties
-   FactSectionAnimals: the data on the number of animals observed in
    each section of each transect
-   FactSectionVegetation: the data on the vegetation in each section
    of each transect
-   FactPlotVegetation: the data on the vegetation in each plot of
    each transect

# 08 Star schema database organization

This example database is organized as a Star Schema database, which is
explained in detail in: 
https://docs.google.com/document/d/1UbUMVFfF4muRqt_YOT73NT7xt-vsoLHs0xNza_Le56U/edit?usp=sharing
This is a particular type of organisation of data into a relational
database using Dim and Fact tables. In addition I also recommend Met
tables. These **Met tables** contain documents, 'data about data'.

-   Dim tables: contain information on lists of objects and subjects
    that you study with their properties
-   Fact tables: are the data that you collect on the objects or
    subjects listed in your Dim tables.

The Dim tables are the points of the star, and the Fact tables are at
the center of the star. The Dim tables are linked to the Fact tables
using the unique ID variable in each Dim table.

# 09 MET tables: metadata

Met tables are a special type of table that contain meta-data, which
is data about the data in your database. Met tables are not part of
the Star Schema, but they are highly recommended to use in your
database. Met tables are used to document the study design, the
variables, and the data collection process. They are also used to
keep track of what still needs to be done in the database, such as
entering data, checking data, etc. Met tables are typically filled in
at the start of the project, and updated during the project.

The following Met tables are recommended to use in your database:

-   MetStudyInfo: short descriptions of the key features of your
    study, such as primary investigator, starting date.
-   MetStill2Do: list of what you still need to do, your ToDo list for
    the database, such as entering data, checking something still in
    your field book, etc. In this table, keep track of who will do it
    of your team, in which table, for which variable and put a date
    when it is complete.
-   MetTables: the list of tables (sheets) in your database, with for
    each table a short contents description and a CSV link to read
    the table in an R script. This link for each table you produce
    from the menu File /Share /Publish to web and then selecting the
    table name (instead of entire document) and Comma separated values
    (instead of Web page). This then shows the link. If you put this
    link in your browser, it gives a download as a csv file. But you
    can also use this link directly in R to read the data using
    readr::read_csv(link). So this avoids the use of intermediary data
    files. Anyone with a script containing that link can read the
    data. Because such links are impossible to guess this is still
    sufficiently safe for regular ecological data. If you however want
    additional security, you can also set up access to tables using
    the google_drive package in R, allowing user authentication.
-   MetVariables: the list of variables in your database, with for
    each variable a short description, the unit, the type of variable
    (numeric, character, factor), and the Dim table it is linked to.
    This is useful to document the variables in your database.

# 10 Dim tables: dimensions

Dim tables contain information on lists of objects and subjects that
you study with their properties. Dim tables are the points of the
star, and the Fact tables are at the center of the star. The Dim
tables are linked to the Fact tables using the unique ID variable in
each Dim table. In this example database, the following Dim tables are
used:

-   DimTransect: the list of transects in the study, with their
    properties such as length, habitat type, etc.
-   DimPlot: the list of plots in the study, with their properties
    such as size, habitat type, etc.
-   DimSpecies: the list of species in the study, with their
    properties such as common name, scientific name, etc.
-   DimSection: the list of sections in the study, with their
    properties such as length, habitat type, etc.

Each Dim table has a unique ID variable that is used to link to the
Fact tables. For example, the DimSpecies table has a unique ID
variable SpeciesID that is used to link to the FactSectionAnimals
table.

# 11 Fact tables: data

Fact tables are the data that you collect on the objects or subjects
listed in your Dim tables. Fact tables are at the center of the star,
and are linked to the Dim tables using the unique ID variable in each
Dim table. In this example database, the following Fact tables are
used:

-   FactSectionAnimals: the data on the number of animals observed in
    each section of each transect, with the SpCode2 variable linking
    to the DimSpecies table and the Section_ID variable linking to the
    DimSection table.
-   FactSectionVegetation: the data on the vegetation in each section
    of each transect, with the Section_ID variable linking to the
    DimSection table.
-   FactPlotVegetation: the data on the vegetation in each plot of
    each transect, with the PlotID linking to the DimPlot table.

Each Fact table has several ID variables that link to the Dim tables,
characterising the observation as which species is observed, at which
transect, etc.

# 12 Reading data from an online database

We will now read in an entire database of transect data with multiple
tables, including metadata. This uses the function read_gsdb that you
defined in the script 01-setup.r. This produces a named list object,
where the elements of the object are different tibbles.

```{r}
transdat<-read_gsdb("https://docs.google.com/spreadsheets/d/1m-liu8omZMewqz_YP9j_YUmQ0zwATl3z4aRLnZFnfWc")
# it contains the following tibbles
names(transdat)
# show the tibble FactSectionAnimals (or explore in the panel Variables)
transdat$FactSectionAnimals
```

# 13 Using ggplot for plotting

ggplot2 is a powerful library allowing all kind of scientific
visualisations/plots.

```{r}
ggplot2::ggplot(data=transdat$FactSectionAnimals, 
                mapping=aes(x=SpCode2,y=CountLeft)) +
  geom_boxplot()
# or use with the same outcome 
ggplot2::ggplot() +
  geom_boxplot(data=transdat$FactSectionAnimals, 
               mapping=aes(x=SpCode2,y=CountLeft))
```

This makes a boxplot of the number of animals observed (CountLeft) for
each species (Spcode6). You can see that the species codes are not
very informative. We can link the species codes to the DimSpecies
table to get the full species names.

```{r}
# the species codes are in the table DimSpecies
transdat$DimSpecies
```

# 14 joining data from different tables

`left_join()` adds columns from a second table while retaining every row
from the table on the left. Start with `FactSectionAnimals`, because it
contains the observations to retain, then join the lookup tables using
their ID variables.

Before joining, check that every key in a lookup table occurs at most
once. Duplicate keys can create extra rows by matching one observation
to multiple records. In this example, each `SpCode2` used in
`FactSectionAnimals` has one matching species record.

```{r}
alldata <- transdat$FactSectionAnimals |>
  dplyr::left_join(transdat$DimSpecies, by = "SpCode2") |>
  dplyr::left_join(transdat$DimSection, by = "Section_ID") |>
  dplyr::left_join(transdat$DimTransect, by = "Transect_ID")

# The number of rows should equal the number of animal observations.
nrow(alldata)
names(alldata)
```

## plot the combined data 
now make the boxplot again, but now with the full species names, 
flip the plot 90 degrees  to make the species names readable, and add axis labels
```{r} 
ggplot2::ggplot(data=alldata, 
                mapping=aes(x=Name_eng,y=CountLeft)) +
  geom_boxplot() +
  labs(x="Species",y="Transect count") +
  coord_flip()
# This makes a boxplot of the number of animals observed (CountLeft) for each species (CommonName)
```
