---
title: "07 Long, wide, and multivariate data"
subtitle: "Reshape wide data into long data using tidyr and ggplot the results"
author: "Han Olff"
date: "2025-9-02"
editor_options: 
  chunk_output_type: console
execute:
  output: false
  warning: false
  message: false
---

# 01 Setup your environment

Run the common setup script and synchronize the local project with
GitHub.

```{r}
# Run the setup script containing package restoration, helper functions, etc.
source(here::here("scripts", "01-setup.R"))

# Pull changes from the online repository into this local project folder.
# This is especially important when collaborating on the same project.
system2("git", c("-C", here::here(), "pull"))

# show which package namespaces are loaded to memory with library()
search()

# authenticate your @student.rug.nl google account
gsheets_auth()
```

# 02 Read and reshape the data

In this script we will work with a vegetation dataset that is in wide format, where each species is a separate column. We will reshape the data into long format, where species is a single variable, and then plot the results.
The input vegetation database can optionally be opened in your browser to inspect it directly.
https://docs.google.com/spreadsheets/d/1WIzkUvunLk1buybZ04Oko8wnF65EbuXjWEviWbeAwu4

Read the vegetation data from the Google Sheet.

```{r}
read_gsdb("https://docs.google.com/spreadsheets/d/1WIzkUvunLk1buybZ04Oko8wnF65EbuXjWEviWbeAwu4", sheets = "FactVegCov", separate = TRUE)
# check which species were recorded in the dataset, and in which years
names(FactVegCov)
unique(FactVegCov$Year)
```

We will now reshape the data using piping from wide to long format, where
'speciesname' will become a single variable instead of distributed over multiple
columns; treat bare, litter, and moss will be treated as if they are species (see the
"data import cheatsheet"). We remove `Salicornia.europaea` and
`Salicornia.procumbens` from the dataset, as `Salicornia.sp` is their
sum of these two; the two species were not separated in earlier years
We will also remove the variables `bare`, `litter`, `mosses` as they are not really species

Reshape the table from wide to long format 
Salicornia species and bare/litter/mosses variables described above
set which original columns to keep (TransectPoint_ID and Year) and which to reshape (all other columns)

```{r}
FactVegCov |> 
  dplyr::select(-c(SalicEur, SalicPro, Bare, Litter, Mosses)) |>
  tidyr::pivot_longer(cols = -c(TransectPoint_ID, Year), 
                      names_to = "speciesname", 
                      values_to = "cover") -> FactVegCov_long

# and show the names of all the species in the new dataset
unique(FactVegCov_long$speciesname) 
```

Now we want to find which species are most common in the dataset. Add a variable to the
dataset that is the rank number of the species according to summed
abundance of each species over the whole dataset (1 = most abundant
species).

```{r}
FactVegCov_long |> 
  dplyr::group_by(speciesname) |> 
  dplyr::summarize(total_cover = sum(cover)) |> 
  dplyr::arrange(desc(total_cover)) |> 
  dplyr::mutate(rank = row_number()) -> FactVegCov_ranked
```

Now merge with the long file, adding a new column that indicates the rank of the species for every observation. This we can use to filter the data for the more common species. 

```{r}
FactVegCov_long |> 
  dplyr::left_join(FactVegCov_ranked, by = "speciesname") -> FactVegCov_long_ranked
print(FactVegCov_long_ranked)
```

# 03 Plot cover of most common species along the transect

Plot the 5 most dominant species as a line diagram, cover (y) versus
`distance_m` (x), with ggplot, a separate plot for each year, each
species with a different line color.

```{r}
FactVegCov_long_ranked |> 
  dplyr::filter(rank <= 5) |> 
  ggplot(aes(x = TransectPoint_ID, y = cover, color = speciesname)) +
  geom_line() +
  facet_wrap(~Year) +
  labs(title = "Cover of Most Common Species Along the Transect",
       x = "Distance (m)",
       y = "Cover (%)")

# save the plot to a png file 
ggsave(filename = here::here("figures/ERS2026-Script07", "ERS2026-Script07-Fig01.png"), 
      width = 1920, height = 1200, 
      units = "px", bg = "white")

```

In this graph it is not easy to see the changes over time
Plot now instead the change in cover along the distance transect, and over the
different years, as a heatmap for the 10 most abundant species (using
ggplot), with a separate heatmap plot per species.

```{r}
FactVegCov_long_ranked |> 
  dplyr::filter(rank <= 10) |> 
  ggplot(aes(x = factor(TransectPoint_ID), y = Year, fill = cover)) +
  geom_tile() +
  facet_wrap(~speciesname) +
  labs(title = "Change in Cover of Most Common Species Along the Transect Over Time",
       x = "Distance (m)",
       y = "Year") +
  scale_fill_viridis_c(option = "plasma", name = "Cover (%)")

# save the plot to a png file
ggsave(filename = here::here("figures/ERS2026-Script07", "ERS2026-Script07-Fig02.png"), 
      width = 1920, height = 1200, 
      units = "px", bg = "white")
```

# 04 Plot change in cover along the elevation gradient

Load the elevation data from 2017-2023, select the variables `Year`,
`TransectPoint_ID`, and `Elevation_m`, and add the `Elevation_m` variable to
the `FactVegCov_long_ranked` vegetation data of 2017-2020.

```{r}
# Load the elevation data
read_gsdb("https://docs.google.com/spreadsheets/d/1UmpZoNIjo5lXdcpGKuBHe7wFX1Z3AQioeHjzrFgYnxo", sheets = "FactElevation", separate = TRUE) 

# Select the required variables
FactElevation |>
  dplyr::select(Year, TransectPoint_ID, Elevation_m) -> FactElevation2

# and join with the vegetation data
FactVegCov_long_ranked |> 
  dplyr::left_join(FactElevation2, by = c("Year", "TransectPoint_ID")) -> FactVegCov_long_ranked2
```

Plot the change in cover along the elevation gradient, and over the
different years, as a heatmap for the 10 most abundant species (using
ggplot), with a separate heatmap plot per species.

```{r}
FactVegCov_long_ranked2 |> 
  dplyr::filter(rank <= 10) |> 
  ggplot(aes(x = Elevation_m, y = cover, fill=speciesname)) +
  geom_area() +
  labs(title = "Change in Cover of 10 Most Common Species Along the Transect Over Time",
       x = "Elevation (m +NAP)",
       y = "cover") +
  facet_wrap(~Year)

# save the plot to a png file
ggsave(filename = here::here("figures/ERS2026-Script07", "ERS2026-Script07-Fig03.png"), 
      width = 1920, height = 1200, 
      units = "px", bg = "white")
```
