---
title: "Big datasets: flooding probability"
subtitle: "Calculate flooding probability at 1 cm resolution from the Schiermonnikoog ferry pier tidal gauge data"
author: "Han Olff"
date: "2025-9-14"
editor_options: 
  chunk_output_type: console
execute:
  output: false
  warning: false
  message: false
---

Data were collected at 10 minute intervals starting 1 January 2020 from the Rijkswaterstaat tidal gauge database `SchierTideDB`
See the database in your browser: 
https://docs.google.com/spreadsheets/d/1DOzvscotzWXm5MmEFrZhvPY820weEZVcSfUYz_5Gyf4
**Warning:** this is a very large database (~2 million records). This
script requires sufficient RAM (at least 16 Gb) and a good processor
(like an i7) to run.

# 01 Setup your environment

Run the common setup script and synchronize the local project with
GitHub.

```{r}
# Run the setup script containing package restoration, helper functions, etc.
source(here::here("scripts", "01-setup.R"))

# Pull changes from the online repository into this local project folder.
# This is especially important when collaborating on the same project.
system2("git", c("-C", here::here(), "pull"))

# show which package namespaces are loaded to memory with library()
search()

# authenticate your @student.rug.nl google account
gsheets_auth()
```

# 02 Read the datasets

Calculate flooding probability from the Rijkswaterstaat tidal gauge
data at the Schiermonnikoog pier, for the Schiermonnikoog tidal data
1980-2025. For working with dates and times, see
[R for Data Science: Dates and Times](https://r4ds.had.co.nz/dates-and-times.html).

```{r}

# Read Database ERS-Schier_tides (takes more than 10 minutes)
read_gsdb("https://docs.google.com/spreadsheets/d/1DOzvscotzWXm5MmEFrZhvPY820weEZVcSfUYz_5Gyf4", 
          sheets="FactSeaLevel", separate=TRUE)
FactSealevel<-read_csv("c://Users/holff/Downloads/factSeaLevel.csv") |>
  dplyr::mutate(DateTime = lubridate::ymd_hms(DateTime)) # convert the DateTime column to a date-time object
tail(FactSealevel, 10) # show the last record of this tiblle
names(FactSealevel)

# Re-format the data for further use.
FactSealevel2 <- FactSealevel |>
  dplyr::mutate(
    Sealevel_cm = ifelse(Sealevel_cm > 1000, NA, Sealevel_cm), # set missing or incorrect values to NA
    Sealevel_m = Sealevel_cm / 100, # cm to m + NAP
    Year = lubridate::year(DateTime),
    Month = lubridate::month(DateTime)
) |>
    dplyr::select(-Sealevel_cm) 
FactSealevel2
max(FactSealevel2$DateTime, na.rm = TRUE) # check the last date in the dataset
```

# 03 Explore tidal level frequency

Show a histogram of tidal levels for summer of each of the last five
years.

```{r}
FactSealevel2 |>
  dplyr::filter(Month >= 4 & Month < 9 & Year %in% 2018:2026 & !is.na(Sealevel_m)) |>
  ggplot(aes(x = Sealevel_m)) + # histogram of tidal heights during this period
  geom_histogram(alpha = 0.5, fill = "blue", binwidth = 0.2) +
  facet_wrap(~Year, nrow = 2) +
  xlab("elevation (m to NAP)") +
  ylab("number of times tide at this level")
```

# 04 Calculate the summer flooding probability

Calculate the April-August flooding frequency of each tidal
elevation, write the results to a file and import in the database `SchierTideDB` as a new table `FloodProb`. The flooding probability is the probability that a particular elevation is flooded at least once during the summer (April-August) of a particular year. The flooding probability is calculated as 1 minus the cumulative frequency of tides at or above that elevation, divided by the total number of tides in that summer.

```{r}
FactSealevel2 |>
  dplyr::filter(Month >= 4 & Month < 8 & !is.na(Sealevel_m) & Year >= 1980) |>
  group_by(Year, Sealevel_m) |>
  summarize(Tides_n = n()) |>
  mutate(
    TransFreq = cumsum(Tides_n),
    Summer_n = max(TransFreq),
    TransProb = round((1 - TransFreq / Summer_n), 3) # transgression probability of each elevation in a year
  ) -> FloodProb
FloodProb
# show the last record of this tiblle
tail(FloodProb, 1)

# Write the result to a csv file. This file goes into the
# `SchierTideDB` database and needs updating every year.

readr::write_csv(FloodProb, here::here("preview", "tempdata", "transgres_freq.csv"))
```

# 05 Explore sea level rise

Show how the summer flooding probability at a particular elevation changes over the years
Before 1990 a different methods was used, so filter those years out

```{r}
FloodProb |>
  dplyr::filter(Sealevel_m == 0.5, Year >= 1990) |> 
  ggplot(aes(x = Year, y = TransProb)) +
  geom_point(size = 3) +
  geom_smooth() +
  xlab("year") +
  ylab("transgression probability of elevation") +
  ggtitle("summer flooding probability at 0.5 m to NAP")
```

The analysis shows that in 2020,2021 and 2022 the salt marsh was exceptionally frequently flooded
while in 2019 is was very little flooded in summer
In general, variability seems to have been higher since 2019 than before


Plot the summer flooding frequency across the elevational gradient,
across all years.

```{r}
FloodProb |> dplyr::filter(Year >= 1990) |>
  ggplot(aes(x = Sealevel_m, y = TransProb, col = factor(Year))) +
  geom_line(linewidth = 1) +
  xlab("elevation (m to NAP)") +
  ylab("flooding probability of elevation")

# plot the same for 2019-2026 only

FloodProb |> dplyr::filter(Year %in% 2019:2026) |>
  ggplot(aes(x = Sealevel_m, y = TransProb, col = factor(Year))) +
  geom_line(linewidth = 1) +
  xlab("elevation (m to NAP)") +
  ylab("flooding probability of elevation")
```
