---
title: "Linear multiple regression with an interaction term"
subtitle: "Modeling clay depth as a function of elevation and year"
editor_options: 
  chunk_output_type: console
execute:
  output: false
  warning: false
  message: false
---

# 01 Setup

Run the common setup script and synchronize the local project with GitHub.

```{r}
# pull the changes in the online repository to this local folder
# this is important when you collaborate on the project
# it updates all your scripts to the latest version, and ensures that you have the same versions of the scripts as your collaborators
system2("git", c("-C", here::here(), "pull"))

# Include the setup script with source(). This will run the code in the setup script, and load the packages and functions that are defined in this script and that we will lead later on in upcoming scripts that we use in this course.
source(here::here("scripts", "01-setup.R"))

# load required libraries for this script
library(lm.beta)

#authenticate with Google Drive to read the dataset from the Google Sheet
googledrive::drive_auth()
```

# 02 Read and explore the dataset
Read the dataset from Google Sheets as a list of tibbles and copy the FactVegClay tibble to a local variable. The dataset contains clay depth measurements from the microtransect in the salt marsh, along with elevation and year of measurement.
```{r}
Microtransect <- read_gsdb("https://docs.google.com/spreadsheets/d/1dJkH09imko9RgOkGzYQT74IeGY56QwiXjjBl7u0KcK0/")
FactVegClay <- Microtransect$FactVegClay
FactVegClay
```

Explore the relation between clay depth and elevation, coloring points by year , and a linear regression line showing how overall ClayDepth_cm changes over with elevation (irrespective of year).

```{r}
FactVegClay |> 
  dplyr::filter(Year<=2023) |>
  ggplot2::ggplot(aes(x = Elevation_m, y = ClayDepth_cm)) +
  geom_point(aes(color = factor(Year))) +
  geom_smooth(method = "lm", se = FALSE)
```

The plot suggests that clay depth decreases with increasing elevation, but also that the relationship between clay depth and elevation may change over time (i.e., the increase seems to be stronger in more recent years). To investigate this further, we will have to fit a linear model with an interaction term between elevation and year.

But first also explore the relation between clay depth and year, coloring points by elevation with a diverging blue-yellow-red scale. 

```{r}
FactVegClay |>
  dplyr::filter(Year<2023) |>
  ggplot2::ggplot(aes(x = Year, y = ClayDepth_cm)) +
  geom_point(aes(color = Elevation_m)) +
  geom_smooth(method = "lm", se = FALSE) +
  geom_smooth(
    method = "lm",
    formula = y ~ x,
    se = FALSE,
    linetype = "dashed"
  ) +
  scale_color_steps2(low = "blue", mid = "yellow", high = "red", 
                     midpoint = mean(FactVegClay$Elevation_m, na.rm = TRUE), n.breaks = 8)
```

The plots suggests that that clay depth tends to increase over time, but the relationship is not consistent across different elevations, at higher elevations the increase seems more pronounced. Again, to investigate this further, we will have to fit a linear model with an interaction term between elevation and year.

# 03 Fit an lm model with interaction term
We will fit a linear model using the with an interaction term between elevation and year, to investigate whether the relationship between clay depth and elevation changes over time.
The model is Elevation_m + Year + Elevation_m:Year, where Elevation_m:Year is the interaction term between elevation and year. The model will be fitted to the data using the lm() function, and the results will be summarized using the broom package.
Note that this model can also be written als ClayDepth_cm ~ Elevation_m * Year, where the * operator indicates that both main effects and the interaction term should be included in the model.
```{r}
FactVegClay |>
  dplyr::filter(Year <= 2023) |>
  stats::lm(ClayDepth_cm ~ Elevation_m + Year + Elevation_m:Year, data = _) -> lm_fit 

# show q-q plot of the residuals to check for normal distribution of residuals, a model assumption of linear regression
broom::augment(lm_fit) |>
  ggplot2::ggplot(aes(sample = .resid)) +
  stat_qq() +
  stat_qq_line() +
  labs(x = "Theoretical quantiles", y = "Sample quantiles") +
  theme_bw()

# summarize different aspects of the fitted model:
  # show the overall fit of the model (R-squared, F-statistic, etc.) 
  broom::glance(lm_fit)
  # show the individual coefficients for the variables in the model and their significance
  broom::tidy(lm_fit)
  # calculate variance inflation factors (VIFs) to check for multicollinearity
  car::vif(lm_fit)
```

The model summary reports several complementary aspects of the fitted model:
1. **Q-q plot of residuals:** It is an assumption of linear regression that the **residuals** are normally distributed. In that case the Q-Q plot should show points approximately along the diagonal line. The Q-Q plot shows in this case that the residuals are indeed approximately normally distributed, with some deviation in the tails. This suggests that the model is a reasonable fit to the data, but there may be some outliers or influential points that could be affecting the fit. A common mistake in this context is that people check if the response variable (y) is normally distributed, but that is not the assumption. It is about the unexplained variables after all the terms in the models have been included. So you can only check this assumption after having fit a model. 

2. **Overall fit:** The $R^2$ value indicates the proportion of variation in `ClayDepth_cm` explained by the predictors. Here, the model explains approximately 55% of the observed variation. However, $R^2$ alone does not establish that the model is appropriate, and a larger value is not automatically preferable if models have different complexity or are being evaluated on different data.

3. **Overall model test:** The F-statistic tests the null hypothesis that all slope coefficients are zero. A very small p-value ($p < 0.001$) provides evidence that the model contains at least one predictor associated with clay depth. It does not show that every predictor is important, nor does it establish causation.

4. **Individual coefficients:** The coefficient table reports estimates, standard errors, t-statistics, and p-values. In this model, `Elevation_m`, `Year`, and the interaction term have p-values below 0.001. Because the model includes a (significant) interaction, the main-effect coefficients should not be interpreted as overall effects: the effect of elevation on claydepth depends on year, and the effect of year on claydepth depends on elevation. These p-values also assume independent errors, an assumption that is revisited later using a mixed-effects model because the same points were measured repeatedly.

5. **Prediction and relative importance:** The fitted equation can be used to obtain predicted clay depth for specified values of the original, uncentered predictors:

    $$
    \text{ClayDepth\_cm} = 3176.8818 - 3636.2477 \times \text{Elevation\_m} - 1.5831 \times \text{Year} + 1.8108 \times \text{Elevation\_m} \times \text{Year}
    $$

    For example, for a point at 1.5 m elevation in 2010, the predicted clay depth is approximately:

    $$
    \text{ClayDepth\_cm} = 3176.8818 - 3636.2477 \times 1.5 - 1.5831 \times 2010 + 1.8108 \times 1.5 \times 2010 \approx 0.17\ \text{cm}.
    $$

    The individual coefficients should not be compared directly as measures of relative importance because they use different measurement scales. For example, changing elevation from metres to centimetres would change its coefficient by a factor of 100 without changing the fitted values or the underlying relationship. Standardized coefficients are more suitable for comparing the slopes on a common scale: they describe the expected change in the response, in standard-deviation units, associated with a one-standard-deviation change in a predictor, conditional on the other terms in the model. However, with an interaction, relative importance still depends on the values of the predictors at which the comparison is made.

    The very large VIF for the interaction term indicates substantial multicollinearity in this uncentered parameterization. Centering the predictors before forming the interaction generally reduces the correlation between the interaction and the main effects, making the coefficients easier to interpret. Centering does not change the fitted values or remove all collinearity. We therefore fit a centered model and calculate standardized coefficients in the next section.


# 04 Fit an lm model with standardized (beta) estimates to assess their relative importance

We will first center  `Elevation_m` and `Year` before fitting, based on the high VIFs calculated. This means that the mean of all elevations is subtracted from each elevation value, and the mean of all years is subtracted from each year value. This centers the data around zero, making the interpretation of the intercept more meaningful and reducing multicollinearity between the main effects and the interaction term. With an interaction term, leaving either predictor uncentered creates severe collinearity between the main effect and the interaction (inflated/misleading coefficients). These original calculated coeffisients are statistically unstable, meaning that a leaving a few observations out of the dataset or including some more would strongly change the value of the coefficients. Centering the predictors reduces this collinearity and stabilizes the coefficients, making them more interpretable.

So we will fit a single model on the centered (but not SD-scaled) predictors, then compute standardized coefficients from that same fit: `lm.beta()` takes the fitted `lm` object and adds a `standardized.coefficients` element to it, rather than requiring a separate refit on rescaled data. So `lm_fit_beta` is not a different model from `lm_fit` above — it *is* the centered `lm` fit, just with standardized coefficients attached, and it still works anywhere an `lm` object would (e.g. when used in `predict()`).

```{r}
FactVegClay |>
    dplyr::filter(Year<=2023) |>
  # center year and elevation
  mutate(
    Elevation_m_c = Elevation_m - mean(Elevation_m),
    Year_c = Year - mean(Year)
  ) |>
  # calculate the multiple regression model with interaction term
  lm(ClayDepth_cm ~ Elevation_m_c * Year_c, data = _) |>
  # calculate the 
  lm.beta::lm.beta() -> lm_fit_beta

# inspect the new standardized coefficients
broom::tidy(lm_fit_beta)
# we do not need to rerun broom::glance(lm_fit_beta) at the standardization does not affect the model fit
```

Interpretation: the standardized coefficients are 0.559 for elevation, 0.503 for year, and 0.146 for the interaction. The elevation and year coefficients indicate comparably sized associations with clay depth when expressed in standard-deviation units, with each predictor considered conditional on the other terms in the model. The smaller interaction coefficient indicates that the change in the elevation–clay-depth relationship across years is more modest on this standardized scale, although the interaction remains statistically significant (see the coefficient table above). The corresponding unstandardized estimates are approximately 24.8 for `Elevation_m_c`, 0.476 for `Year_c`, and 2.53 for the interaction. Centering changes the interpretation of the coefficients and reduces parameterization-related collinearity; calculating standardized coefficients changes their scale. Neither operation changes the fitted values, overall model fit, or p-values for the corresponding terms.

# 05 plot the fitted model 

A naive way of proceeding would be to plot the model in ggplot for elevation versus clay-depths, with geom_smooth(method = "lm") combined year as a color aesthetic in the aes(). Shows that the effect of elevation is different for different years, sugggesting the interaction that we found in the model. 

```{r}
FactVegClay |>
    dplyr::filter(Year<=2023) |>
  ggplot2::ggplot(aes(x = Elevation_m, y = ClayDepth_cm, color = factor(Year))) +
  geom_point() +
  geom_smooth(method = "lm", se = FALSE) +
  labs(x = "Elevation (m)", y = "Clay depth (cm)", color = "Year") +
  labs(title = "Fitted values are not from the year x elevation interaction model") +
  theme_bw()
```

But note that we are not the  model here that we just fitted. Instead we calculate here  a separate lm fit for each year! It does not accomodate year as having a linear effect on claydepth, as we tested in our model. To plot the fitted values from our model, we can use geom_line() with the fitted values from the centered lm fit `lm_fit_beta$fitted.values`. This works again because of row-order correspondence, not because the model is somehow evaluating uncentered predictors. `lm_fit_beta$fitted.values` is a vector of the already-computed fitted values from fitting on the centered data — one number per row of the training data `FactVegClay`, in the same row order.

```{r}
# plot the model with the fitted values from the centered lm fit
FactVegClay |>
    dplyr::filter(Year<=2023) |>
  ggplot2::ggplot(aes(x = Elevation_m, y = ClayDepth_cm, color = factor(Year))) +
  geom_point() +
  geom_line(aes(y = lm_fit_beta$fitted.values), linewidth = 1) +
  labs(x = "Elevation (m)", y = "Clay depth (cm)", color = "Year") +
  theme_bw()

# predict clay depth at 1.05 m and 1.2 m elevation in 2017  and year 2025
predict(
  lm_fit_beta,
  newdata = data.frame(
    Elevation_m_c = c(1.05 - mean(FactVegClay$Elevation_m), 1.2 - mean(FactVegClay$Elevation_m)),
    Year_c = c(2017 - mean(FactVegClay$Year), 2025 - mean(FactVegClay$Year)),
    row.names = c("1.05", "1.2")
  )
)
```

Our final interpretation of the analysis is that the predicted sedimentation rate (increase in clay depth over time) is higher at higher elevations than at lower elevations. At 1.05 m elevation, the model predicts an increase in clay depth from approximately 0.83 cm in 2017 to 2.73 cm in 2023, an increase of 1.91 cm over six years. At 1.20 m elevation, it predicts an increase from approximately 3.25 cm to 6.79 cm, an increase of 3.54 cm over the same period. This pattern is consistent with some factor inhibiting sedimentation at lower elevations. Possible explanations include sparse vegetation associated with prolonged waterlogging, which may lead to anoxia and high sulfide concentrations, as well as the removal of sediment when microbial mats dry out and are washed away. 

# 06 Accounting for repeated measures on the same transect points

`FactVegClay` (filtered to `Year <= 2023` in this section) is a balanced panel: 71 `Point_ID`s, each measured in every one of 8 years, 2017-2023 (497 = 71 × 7 rows -- see `broom::glance()` above for the exact `nobs`). Observations from the same point across years are not independent, so the `lm()` models above understate the uncertainty of the fixed-effect estimates. We fit a mixed-effects model with `glmmTMB`, adding a per-point random intercept and random slope for `Elevation_m_c` (since `Elevation_m` changes over time within a point, its effect may vary by point too).

We considered adding an AR(1) residual correlation structure across years within each point (`glmm_fit_ar1`, using `ar1(Year_f + 0 | Point_ID)` with `Year` coded as a factor `Year_f`) to capture genuine temporal autocorrelation, beyond the general non-independence a random intercept/slope already accounts for. That combined model produces a convergence warning (non-positive-definite Hessian) and the residual variance collapses to near zero — with only 71 points measured a handful of times each, there isn't enough replication to identify both a per-point random-slope covariance structure and a separate AR(1) term at once. We therefore drop the AR(1) term and instead compare two random-slope structures, both fit with a Gaussian family and identity link: one letting the per-point random intercept and slope correlate (`(Elevation_m_c | Point_ID)`, `glmm_fit_corr`), and a simpler one that fixes that correlation to zero (`(Elevation_m_c || Point_ID)`, `glmm_fit`). Both converge cleanly on their own (no AR(1) term); a likelihood-ratio test (`anova(glmm_fit, glmm_fit_corr)`) shows the intercept-slope correlation does not significantly improve the fit, so we prefer the simpler, uncorrelated model on parsimony grounds and inspect it with `summary()`.

```{r}
library(glmmTMB)

FactVegClay_glmm <- FactVegClay |>
  dplyr::filter(Year <= 2023) |>
  mutate(
    Elevation_m_c = Elevation_m - mean(Elevation_m),
    Year_c = Year - mean(Year),
    Year_f = factor(Year)
  )

# Random slope + AR(1) residual correlation 
glmm_fit_ar1 <- glmmTMB(
  ClayDepth_cm ~ Elevation_m_c * Year_c +
    (Elevation_m_c || Point_ID) +
    ar1(Year_f + 0 | Point_ID),
  family = gaussian("identity"),
  data = FactVegClay_glmm
)
summary(glmm_fit_ar1)
# But this does not converge reliably
# (non-positive-definite Hessian, residual variance collapses to ~0) --
# too many random-effect parameters for 71 points x 9 years of data.

# correlated random intercept and slope
glmm_fit_corr <- glmmTMB(
  ClayDepth_cm ~ Elevation_m_c * Year_c + (Elevation_m_c | Point_ID),
  family = gaussian("identity"),
  data = FactVegClay_glmm
)
broom.mixed::tidy(glmm_fit_corr)
broom.mixed::glance(glmm_fit_corr)

# uncorrelated random intercept and slope
glmm_fit <- glmmTMB(
  ClayDepth_cm ~ Elevation_m_c * Year_c + (Elevation_m_c || Point_ID),
  family = gaussian("identity"),
  data = FactVegClay_glmm
)
broom.mixed::tidy(glmm_fit)
broom.mixed::glance(glmm_fit)

# LRT: does allowing the intercept-slope correlation improve the fit?
anova(glmm_fit, glmm_fit_corr)

summary(glmm_fit)
```

Conclusion: repeated measures on the 71 transect points across 8 years (2017–2023 in this filtered dataset) mean observations aren't independent, so a plain `lm()` understates uncertainty. Three `glmmTMB` mixed models were compared: a random-slope-plus-AR(1) model (`glmm_fit_ar1`) failed to converge reliably (non-positive-definite Hessian, residual variance collapsing toward zero) since the data can't identify both a per-point slope covariance structure and a separate AR(1) term simultaneously. Of the two remaining models — with and without a correlation between the per-point random intercept and slope — a likelihood-ratio test (χ²(1) = 0.52, p = 0.47) found the correlation didn't significantly improve fit, so the simpler uncorrelated model (`glmm_fit`) was preferred. Its fixed-effect estimates (elevation ≈ 21.4, year ≈ 0.48, interaction ≈ 1.73, all significant) closely match the earlier `lm_fit_beta` estimates, but the standard errors here properly account for the repeated-measures structure rather than treating all rows as independent.