---
title: "10 generalized linear models (GLMs)"
subtitle: "Model plant occurrence along an elevation gradient with logistic regression"
author: "Han Olff"
date: "2025-09-16"
editor_options:
  chunk_output_type: console
execute:
  output: false
  warning: false
  message: false
---
# 00 Introduction to Generalized Linear Models 
In Generalized Linear Models (GLMs) we let go of the assumption of Linear Models (LMs) that the residuals need to be normally distributed. Instead they can follow a binomal distribution, poission distribution, negative binomial distribution, or various zero-inflated distributions. 
Typically the type of regression you do in a GLM is named after the distribution. For example a logistic regression (bionomial distribution), poisson regression (poisson distribution), etc. 

You can use the following decision tree for determining the best GLM for your data, depending on a series of questions

What kind of Y do I have?

Continuous
├── Can be negative / approximately symmetric
│      → Gaussian
└── Non-negative
       ├── No zeros
       │      → Gamma
       └── Exact zeros present
              → Tweedie
              OR hurdle model if zero and positive values
                 represent different biological processes

Counts
├── Variance ≈ mean
│      → Poisson
├── Variance > mean
│      → Negative binomial
└── More zeros than the count model predicts
       → Zero-inflated Poisson
       → Zero-inflated negative binomial
       → or hurdle model
Binary
    → Binomial

Proportion
├── successes / number of trials
│      → Binomial
└── continuous proportion
       → Beta regression


This exercise uses logistic regression to model the presence (1) or absence
(0) of salt-marsh plant species (so binomial response) along a microtransect. We compare linear and quadratic elevation responses, make predictions, and visualise the fitted
occurrence probabilities.

# 01 Setup your environment
we will load the libraries glmmTMB for fitting GLMs and DHARMa for testing assumptions, as degree of zero-inflation etd
```{r}
# Run the common setup script containing packages and helper functions.
source(here::here("scripts", "01-setup.R"))

# Authenticate your @student.rug.nl Google account.
gsheets_auth()

library(glmmTMB) # for fitting generalized linear models (GLMs)
library(DHARMa)  # for checking assumptions of GLMs
```

# 02 Read and prepare the data

The `FactVegClay` table stores occurrence per species in wide format. Read the
microtransect database, reshape the species columns to long format, and retain
the 2025 observations (except `Salicornia.sp`).

```{r}
# Database: MicrotransectDB
MicrotransectDB <- read_gsdb(
  "https://docs.google.com/spreadsheets/d/1dJkH09imko9RgOkGzYQT74IeGY56QwiXjjBl7u0KcK0/"
)
FactVegClay2026_wide<-MicrotransectDB$FactVegClay |>
  dplyr::filter(Year==2026)
FactVegClay2026_long <- MicrotransectDB$FactVegClay |>
  tidyr::pivot_longer(
    cols = -c(Point_ID:ClayDepth_cm),
    names_to = "Species",
    values_to = "Present"
  ) |>
  dplyr::arrange(Year, Species, Point_ID) |>
  dplyr::filter(Year == 2026, Species != "Salicornia.sp")

FactVegClay2026_long
```

Calculate the thickness of the sand and clay layers and display their depth
along the transect.

```{r}
Soil2026 <- FactVegClay2026_wide |>
  dplyr::mutate(SandDepth_cm = 100 * Elevation_m - ClayDepth_cm) |>
  dplyr::select(Point_ID, SandDepth_cm, ClayDepth_cm) |>
  tidyr::pivot_longer(
    cols = -Point_ID,
    names_to = "SoilLayer",
    values_to = "LayerThickness_cm"
  )

p1 <-Soil2026 |>
  ggplot(aes(x = Point_ID, y = LayerThickness_cm, fill = SoilLayer)) +
  geom_area() +
  coord_cartesian(ylim = c(95, 125)) +
  scale_fill_manual(values = c(
    "SandDepth_cm" = "khaki",
    "ClayDepth_cm" = "saddlebrown"
  )) +
  labs(x = "Distance along transect (m)",
   y = "Layer thickness (cm)",
   title="clay profile")
p1
```

Show the observed occurrences of all species, then combine this figure with
the soil profile.

```{r}
PlantPres <- FactVegClay2026_long |>
  dplyr::filter(Present == 1) |>
  dplyr::mutate(SpeciesNum = as.numeric(factor(Species)))

p2 <- ggplot(
  PlantPres,
  aes(x = Point_ID, y = max(SpeciesNum) - SpeciesNum, colour = Species)
) +
  geom_point(size = 2) +
  labs(x = NULL, y = 'species', title='species occurrence') +
  theme(axis.text.x = element_blank(), axis.text.y = element_blank())
p2
# combine the ggplot objects in one plot with the patchwork package
p2 / p1
```

# 03 Explore species responses to elevation

Calculate the frequency of occurrence of each species in 2025 and select the
six most common species for modelling.

```{r}
SpeciesFreq2026 <- FactVegClay2026 |>
  dplyr::group_by(Species) |>
  dplyr::summarise(Freq = mean(Present), .groups = "drop") |>
  dplyr::arrange(desc(Freq))
SpeciesFreq2026

DomSpec <- c(
  "Puccinellia.maritima", "Salicornia.europaea", "Limonium.vulgare",
  "Plantago.maritima", "Glaux.maritima", "Salicornia.procumbens"
)
DomSpec
```

# 04 Fit logistic regression models

For a binary response, logistic regression models the log odds of occurrence.
First fit linear and quadratic elevation responses for *Glaux maritima*.

```{r}
m.Glaux1 <- MicrotransectDB$FactVegClay2026 |>
  dplyr::filter(Species == "Glaux.maritima") |>
  glmmTMB::glmmTMB(
    Present ~ Elevation_m,
    family = binomial(link = "logit"),
    data = _
  )
summary(m.Glaux1)

m.Glaux2 <- MicrotransectDB$FactVegClay2025 |>
  dplyr::filter(Species == "Glaux.maritima") |>
  glmmTMB::glmmTMB(
    Present ~ Elevation_m + I(Elevation_m^2),
    family = binomial(link = "logit"),
    data = _
  )
summary(m.Glaux2)

# Compare the two models. A lower AIC (by about 2 or more) favours that model.
AIC(m.Glaux1, m.Glaux2)
```

Define reusable functions to fit each candidate model and add response-scale
predictions to the data. The predictions are probabilities of occurrence.

```{r}
fit1 <- function(species_name, db = FactVegClay2025) {
  db |>
    dplyr::filter(Species == species_name) |>
    glmmTMB::glmmTMB(
      Present ~ Elevation_m,
      family = binomial(link = "logit"),
      data = _
    )
}

fit2 <- function(species_name, db = FactVegClay2025) {
  db |>
    dplyr::filter(Species == species_name) |>
    glmmTMB::glmmTMB(
      Present ~ Elevation_m + I(Elevation_m^2),
      family = binomial(link = "logit"),
      data = _
    )
}

add_predictions <- function(species_name, model,
                            db = FactVegClay2025) {
  predictions <- if ("Predicted" %in% names(db)) {
    db$Predicted
  } else {
    rep(NA_real_, nrow(db))
  }

  rows <- db$Species == species_name
  predictions[rows] <- predict(
    model,
    newdata = db[rows, , drop = FALSE],
    type = "response"
  )

  db$Predicted <- predictions
  db
}
```

Inspect the two candidate models for each dominant species. Select a model
using the summaries and AIC comparison, then add its predictions to the
dataset.

```{r}
# Glaux maritima: linear model
summary(fit1("Glaux.maritima"))
summary(fit2("Glaux.maritima"))
FactVegClay2025 <- add_predictions(
  "Glaux.maritima", fit1("Glaux.maritima")
)

# Limonium vulgare: linear model
summary(fit1("Limonium.vulgare"))
summary(fit2("Limonium.vulgare"))
FactVegClay2025 <- add_predictions(
  "Limonium.vulgare", fit1("Limonium.vulgare")
)

# Plantago maritima: quadratic model
summary(fit1("Plantago.maritima"))
summary(fit2("Plantago.maritima"))
FactVegClay2025 <- add_predictions(
  "Plantago.maritima", fit2("Plantago.maritima")
)

# Puccinellia maritima: linear model
summary(fit1("Puccinellia.maritima"))
summary(fit2("Puccinellia.maritima"))
FactVegClay2025 <- add_predictions(
  "Puccinellia.maritima", fit1("Puccinellia.maritima")
)

# Salicornia europaea: quadratic model
summary(fit1("Salicornia.europaea"))
summary(fit2("Salicornia.europaea"))
FactVegClay2025 <- add_predictions(
  "Salicornia.europaea", fit2("Salicornia.europaea")
)

# Salicornia procumbens: linear model
summary(fit1("Salicornia.procumbens"))
summary(fit2("Salicornia.procumbens"))
FactVegClay2025 <- add_predictions(
  "Salicornia.procumbens", fit1("Salicornia.procumbens")
)
```

We can use the package DHARMa  for model diagnostics (it is not for testing  predictor significance)
For example, check dispersion, zero inflation, and outliers for one selected
model. 

```{r}
glaux_residuals <- DHARMa::simulateResiduals(fit1("Glaux.maritima"), n = 1000)
plot(glaux_residuals)
DHARMa::testDispersion(glaux_residuals)
DHARMa::testZeroInflation(glaux_residuals)
DHARMa::testOutliers(glaux_residuals)
```

Because for the dispersion test, $P > 0.05$, the test is not significant, so there is no evidence of overdispersion or underdispersion in this model. This is the desirable result for this diagnostic
DHARMa::testZeroInflation(glaux_residuals) compares the number of observed zeros with the number expected under the fitted model. in this case, The ratio is almost exactly 1, meaning the model generated approximately the expected number of zeros. Since $p = 1$, there is no evidence of excess zero inflation in the Glaux maritima model.

# 05 Plot the fitted occurrence probabilities

```{r}
# Observed presences/absences and fitted probabilities per species.
p4 <- FactVegClay2025 |>
  dplyr::filter(Species %in% DomSpec) |>
  ggplot(aes(x = Elevation_m, y = Present)) +
  geom_point(size = 1) +
  geom_line(aes(y = Predicted)) +
  labs(title = "Microtransect", x = "Elevation (m)", y = "Presence") +
  ylim(-1, 2) +
  facet_wrap(~Species)
p4

# Fitted occurrence probabilities for all six species in one plot.
p5 <- FactVegClay2025 |>
  dplyr::filter(Species %in% DomSpec) |>
  ggplot(aes(x = Elevation_m, colour = Species)) +
  geom_line(aes(y = Predicted), linewidth = 2) +
  labs(title = "Microtransect", x = "Elevation (m)", y = "Presence") +
  ylim(0, 1)
p5

# Combine the final two plots vertically.
p4 / p5
```
