---
title: "General linear models (GLM) - Poisson regression"
subtitle: "find the best model for the relation between cockle counts and elevation"
author: "Han Olff"
date: "2025-9-02"
editor_options: 
  chunk_output_type: console
execute:
  output: false
  warning: false
  message: false
---

# 01 Setup your environment

Run the common setup script and synchronize the local project with GitHub.

```{r}
# Run the setup script containing package restoration, helper functions, etc.
source(here::here("scripts", "01-setup.R"))

# Pull changes from the online repository into this local project folder.
# This is especially important when collaborating on the same project.
system2("git", c("-C", here::here(), "pull"))

# show which package namespaces are loaded to memory with library()
search()

# authenticate your @student.rug.nl google account
gsheets_auth()
```


# 02 Read and combine the data

Load the elevation data and inspect its structure. Extract the data for 2026 into a new tibble, keep only the variables `distance_m` and `elevation`, and omit records where `TransectPoint_ID` is missing (`NA`).

```{r}
elevdata<-read_gsdb("https://docs.google.com/spreadsheets/d/1UmpZoNIjo5lXdcpGKuBHe7wFX1Z3AQioeHjzrFgYnxo")
FactElevation2026<-elevdata$FactElevation |>
  dplyr::filter(Year == 2026, !is.na(TransectPoint_ID)) |>
  dplyr::select(TransectPoint_ID, DistanceRTK_m, Elevation_m)
```



Read the cockle data: keep only the data for 2026

```{r}
# read the cockle data
cockledata<-read_gsdb("https://docs.google.com/spreadsheets/d/1E1vUWAHhse7fhjBf94Kiog3Rsj7IkMuFtP-BTQX2oI8")

FactCockleSize2026 <- cockledata$FactCockleSize |>
  dplyr::filter(Year == 2026, !is.na(TransectPoint_ID)) |>
  dplyr::select(Year, ReplicatePlot_ID, CockleObs_ID, TransectPoint_ID, Length_mm) |>
  dplyr::group_by(TransectPoint_ID) |>
  dplyr::summarise(
    n_obs = n(),
    avg_l = mean(Length_mm, na.rm = T),
    sd_l = sd(Length_mm, na.rm = T), # measure of group variation
    se_l = sd_l / sqrt(n_obs) # measure of how precise you measure the mean
  )
print(FactCockleSize2026)
```

Plot (with a line and points) how the number of cockles changes with distance (`TransectPoint_ID`) along the transect.

```{r}
FactCockleSize2017 |>
  ggplot2::ggplot(aes(x = TransectPoint_ID, y = n_obs)) +
  geom_point() +
  geom_line() +
  labs(x = "transect point", y = "number of cockles per 3 plots")
```

Merge the cockle and elevation data into a single table, `combidat`, using `TransectPoint_ID` as the common variable between the two tables, and replace all `NA` in `n_obs` with zero.

```{r}
Combidat <-
  dplyr::left_join(FactElevation2017, FactCockleSize2017,
    by = "TransectPoint_ID"
  ) |>
  dplyr::mutate(n_obs = tidyr::replace_na(n_obs, 0)) # replace na by 0

Combidat
```

we find that the number of cockles is highest in the midde of this part of the transect, between transectpoint 170 and 2000

# 04 Cockle density and elevation

Explore how the cockle density changes with elevation.

```{r}
Combidat |>
  ggplot2::ggplot(aes(x = Elevation_m, y = n_obs)) +
  geom_point(size = 2) +
  geom_line()

# Fit a linear regression: `~` means "depends on / as a function of", so `lm(response_variable ~ predictor_variable)`.
linreg <- stats::lm(n_obs ~ Elevation_m, data = Combidat)
summary(linreg)

# Show this model as a line in ggplot, with the confidence interval.

Combidat |>
  ggplot2::ggplot(aes(x = Elevation_m, y = n_obs)) +
  geom_point(size = 2) +
  geom_smooth(method = "lm", formula = "y~x")

# this is clearly not a good model, as the relation is clearly not linear, but rather unimodal. So we will fit a better model, using a loess smoother, and show this model in ggplot.
Combidat |>
  ggplot2::ggplot(aes(x = Elevation_m, y = n_obs)) +
  geom_point(size = 2) +
  geom_smooth(method = "loess", fill = NA) +
  labs(x = "elevation in m + NAP", y = "number of cockles per 3 plots")
```

# 05 Cockle size and elevation

Plot how the size (as mean length) of cockles changes with elevation along the transect. Omit the observations where length is `NA` (because no cockles were measured), fit a quadratic model (second-order polynomial), show for each point also the standard errors, and add appropriate labels for the x and y axis.

```{r}
Combidat |>
  dplyr::filter(!is.na(avg_l)) |>
  ggplot2::ggplot(aes(x = Elevation_m, y = avg_l)) +
  geom_point() +
  geom_smooth(method = "lm", 
              formula = "y~x",
              fill = NA) +
  geom_smooth(method = "lm",
              formula = "y~x+I(x^2)",
              se = TRUE,
              color = "red", fill = NA) +
  xlim(0, 0.8) +
  labs(x = "elevation in m + NAP", y = "mean cockle length (mm)")

# Test with linear regression whether these relations are significant.

linreg2 <- stats::lm(avg_l ~ Elevation_m, data = Combidat)
summary(linreg2)
linreg3 <- stats::lm(avg_l ~ Elevation_m + I(Elevation_m^2), data = Combidat)
summary(linreg3)
```

The unimodal relation between cockle size and elevation is not significant, as the quadratic term is not significant. The linear relation between cockle size and elevation is significant, as the p-value of the linear term is \< 0.05. We conclude that bigger cockles just found lower on the transect, without evidence for an optimum.