---
title: "General linear models (GLM) - Poisson regression"
subtitle: "find the best model for the relation between cockle counts and elevation"
author: "Han Olff"
date: "2025-9-02"
editor_options: 
  chunk_output_type: console
execute:
  output: false
  warning: false
  message: false
---

# 01 Setup your environment

Run the common setup script and synchronize the local project with GitHub.

```{r}
# Run the setup script containing package restoration, helper functions, etc.
source(here::here("scripts", "01-setup.R"))

# Pull changes from the online repository into this local project folder.
# This is especially important when collaborating on the same project.
system2("git", c("-C", here::here(), "pull"))

# show which package namespaces are loaded to memory with library()
search()

# authenticate your @student.rug.nl google account
gsheets_auth()
```

# 02 Read and combine the data

Load the elevation data and extract the data for 2026 into a new tibble, keep only the variables `distance_m` and `elevation`, and omit records where `TransectPoint_ID` is missing (`NA`).

```{r}
elevdata<-read_gsdb("https://docs.google.com/spreadsheets/d/1UmpZoNIjo5lXdcpGKuBHe7wFX1Z3AQioeHjzrFgYnxo")
FactElevation2026<-elevdata$FactElevation |>
  dplyr::filter(Year == 2026, !is.na(TransectPoint_ID)) |>
  dplyr::select(TransectPoint_ID, DistanceRTK_m, Elevation_m)
```

Read the cockle data: keep only the data for 2026 and define three size cohorts

```{r}
# read the cockle data
cockledata<-read_gsdb("https://docs.google.com/spreadsheets/d/1E1vUWAHhse7fhjBf94Kiog3Rsj7IkMuFtP-BTQX2oI8")

# histogram of the size of cockles in 2026
cockledata$FactCockleSize |>
  dplyr::filter(Year == 2026, !is.na(TransectPoint_ID),Present==1) |>
  ggplot2::ggplot(aes(x = Length_mm)) +
  geom_histogram(binwidth = 1) +
  labs(x = "length of cockles (mm)", y = "number of cockles")

# define three cohorts (small, medium, large) based on the length of cockles in 2026, representing likely age classes (1, 2, 3 and more years old). We will use these cohorts later to explore whether the relation between cockle size and elevation differs between cohorts.
cockledata$FactCockleSize <- cockledata$FactCockleSize |>
  dplyr::mutate(
    cohort = dplyr::case_when(
      Length_mm < 15 ~ "young",
      Length_mm >= 15 & Length_mm <23 ~ "medium",
      Length_mm >= 23 ~ "old"
    ),
    cohort = factor(cohort, levels = c("young", "medium", "old")) # set in right order instead of alphabetical order
  )

# histogram of the size of cockles in 2026 again now with cohorts
cockledata$FactCockleSize |>
  dplyr::filter(Year == 2026, !is.na(TransectPoint_ID),Present==1) |>
  ggplot2::ggplot(aes(x = Length_mm, fill=cohort)) +
  geom_histogram(binwidth = 1, boundary=0) +
  labs(x = "length of cockles (mm)", y = "number of cockles")


#Make a tibble with all combinations of transectpoints and replicate plots assessed for cockles in 2026, so that we can also include the transectpoints where no cockles were found in 2026.
design2026<-expand.grid(ReplicatePlot_ID=unique(cockledata$FactCockleSize$ReplicatePlot_ID[cockledata$FactCockleSize$Year == 2026]),
                        TransectPoint_ID=unique(cockledata$FactCockleSize$TransectPoint_ID[cockledata$FactCockleSize$Year == 2026])) |>
  dplyr::as_tibble()
design2026


# sum up  the number of cockles per transectpoint per replicate plot in 2026.
CockleCount2026present <- cockledata$FactCockleSize |>
  dplyr::filter(Year == 2026, !is.na(TransectPoint_ID),Present==1) |>
    dplyr::summarise(.by = c("TransectPoint_ID", "ReplicatePlot_ID", "cohort"), cockle_count = dplyr::n())
CockleCount2026present

# merge the design with the observation data, so that we also have the transect points where no cockles were found in 2026, and replace all `NA` in `n_obs` with zero, and also merge with the elevation data for 2026
CockleCount2026 <- dplyr::left_join(design2026, CockleCount2026present, by = c("TransectPoint_ID", "ReplicatePlot_ID")) |>
    dplyr::left_join(FactElevation2026, by = "TransectPoint_ID") |>
  dplyr::mutate(cockle_count = tidyr::replace_na(cockle_count, 0)) # replace na by 0

print(CockleCount2026)
```

Plot with a linear model how the number of cockles per transectpoint changes with elevation, all fonts size 10

```{r}
CockleCount2026 |> 
  dplyr::filter(Elevation_m > 0.25) |>  #remove outliers with very low elevation, not representative for the rest of the transect 
  ggplot2::ggplot(aes(x = Elevation_m, y = cockle_count)) +
  geom_point(aes(color = factor(ReplicatePlot_ID)), size = 1.5, alpha = 0.5) +
  geom_smooth(method = "lm", formula = "y~x") +
  labs(
    x = "Elevation (m + NAP)",
    y = "number of cockles per 0.068 m2",
    color = "replicate plot"
  ) +
  theme(text = element_text(size = 10))
```

Note the large variation between the replicate plots#

# 04 Model the relation between cockle count and elevation

Explore how the cockle density changes with elevation.

```{r}
# Fit a linear regression: 
linreg <- stats::lm(cockle_count ~ Elevation_m, data = CockleCount2026)
summary(linreg)

# The model is not significant. We would conclude that there is no evidence for a linear relation between cockle count and elevation.
# but check the q-q plot of the residuals to see whether the assumptions of the linear model are met (Gaussian distribution of residuals)
ggplot2::ggplot(linreg, aes(sample = .stdresid)) +
  geom_qq() +
  geom_qq_line() +
  labs(x = "theoretical quantiles", y = "standardized residuals") +
  theme(text = element_text(size = 10))
# to test if there is a significant deviation from a normal distribution: 
stats::shapiro.test(residuals(linreg))


# this suggestes that the residuals are not at all normally distributed, and therefore the assumptions of the linear model are not met. We will therefore fit a generalized linear model (GLM) with a Poisson distribution, which is more appropriate for count data.

CockleCount2026 |> 
  dplyr::filter(Elevation_m > 0.25) |>  #remove the outlier transect+id with very low elevation, not representative 
  glm(cockle_count ~ Elevation_m, data = _, family = "poisson") -> glm1
summary(glm1) 

# check if the model improves if we add a quadratic term (second-order polynomial) to the model, to test for a unimodal relation between cockle count and elevation.
CockleCount2026 |> 
  dplyr::filter(Elevation_m > 0.25) |>  #remove the outlier transect+id with very low elevation, not representative 
  glm(cockle_count ~ Elevation_m + I(Elevation_m^2), data = _, family = "poisson") -> glm2
summary(glm2) 

# yes the quadratic term is significant, so we conclude that there is a unimodal relation between cockle count and elevation. Also the AIC is lower for the quadratic model, indicating a better fit. We will therefore use this model to predict the number of cockles per transectpoint as a function of elevation.

# check the q-q plot of the residuals to see whether the assumptions of the GLM are met, using simulated residuals from the DHARMa package, which is more appropriate for GLMs than the standard residuals.
library(DHARMa)
simres <- DHARMa::simulateResiduals(fittedModel = glm2, n = 1000)
plot(simres)
# this doesn't look good, the residuals are not uniformly distributed, and there is a clear pattern in the residuals, indicating that the model does not fit the data well. We will therefore test for uniformity and dispersion of the residuals using the DHARMa package.

# test for uniformity and dispersion of the residuals. A Poisson distribution assumes that the variance is equal to the mean, and therefore the dispersion parameter should be close to 1. If the dispersion parameter is larger than 1, this indicates overdispersion, which means that the variance is larger than the mean, and therefore the Poisson distribution is not appropriate for these data. In that case we can try to fit a negative binomial model instead of a Poisson model.
DHARMa::testDispersion(simres)



CockleCount2026 |> 
  dplyr::filter(Elevation_m > 0.25) |>  #remove the outlier transect+id with very low elevation, not representative 
  ggplot2::ggplot(aes(x = Elevation_m, y = cockle_count)) +
  geom_point(aes(color = factor(ReplicatePlot_ID)), size = 1.5, alpha = 0.7) +
  geom_smooth(method = "glm", formula = "y~x+I(x^2)", method.args = list(family = "poisson")) +
  labs(
    x = "Elevation (m + NAP)",
    y = "number of cockles per 0.068 m2",
    color = "replicate plot"
  ) +
  theme(text = element_text(size = 10))

```

we find that the number of cockles is highest in the middle of this part of the transect, between transectpoint 170 and 2000, and that the relation between cockle count and elevation is unimodal. The only problem that we have not addresses is that the data are not independent, as the same transectpoint is measured in three replicate plots, that are assumed to have the same elevation. We will address this issue in a next script, by fitting a generalized linear mixed model (GLMM) with transectpoint as a random effect.