---
title: "General linear models (GLM) - Poisson regression"
subtitle: "find the best model for the relation between cockle counts and elevation"
author: "Han Olff"
date: "2025-9-02"
editor_options: 
  chunk_output_type: console
execute:
  output: false
  warning: false
  message: false
---

# 01 Setup your environment

Run the common setup script and synchronize the local project with GitHub.

```{r}
# Run the setup script containing package restoration, helper functions, etc.
source(here::here("scripts", "01-setup.R"))

# Pull changes from the online repository into this local project folder.
# This is especially important when collaborating on the same project.
system2("git", c("-C", here::here(), "pull"))

# show which package namespaces are loaded to memory with library()
search()

# authenticate your @student.rug.nl google account
gsheets_auth()

# load libraries used in this script
library(DHARMa)

```

# 02 Read and combine the data

Load the elevation data and extract the data for 2026 into a new tibble, keep only the variables `distance_m` and `elevation`, and omit records where `TransectPoint_ID` is missing (`NA`).

```{r}
elevdata<-read_gsdb("https://docs.google.com/spreadsheets/d/1UmpZoNIjo5lXdcpGKuBHe7wFX1Z3AQioeHjzrFgYnxo")
FactElevation2026<-elevdata$FactElevation |>
  dplyr::filter(Year == 2026, !is.na(TransectPoint_ID)) |>
  dplyr::select(TransectPoint_ID, DistanceRTK_m, Elevation_m)
```

Read the cockle data: keep only the data for 2026 and define three size cohorts

```{r}
# read the cockle data
cockledata<-read_gsdb("https://docs.google.com/spreadsheets/d/1E1vUWAHhse7fhjBf94Kiog3Rsj7IkMuFtP-BTQX2oI8")

# histogram of the size of cockles in 2026
cockledata$FactCockleSize |>
  dplyr::filter(Year == 2026, !is.na(TransectPoint_ID),Present==1) |>
  ggplot2::ggplot(aes(x = Length_mm)) +
  geom_histogram(binwidth = 1) +
  labs(x = "length of cockles (mm)", y = "number of cockles")

# define three cohorts (small, medium, large) based on the length of cockles in 2026, representing likely age classes (1, 2, 3 and more years old). We will use these cohorts later to explore whether the relation between cockle size and elevation differs between cohorts.
cockledata$FactCockleSize <- cockledata$FactCockleSize |>
  dplyr::mutate(
    cohort = dplyr::case_when(
      Length_mm < 15 ~ "young",
      Length_mm >= 15 & Length_mm <23 ~ "medium",
      Length_mm >= 23 ~ "old"
    ),
    cohort = factor(cohort, levels = c("young", "medium", "old")) # set in right order instead of alphabetical order
  )

# histogram of the size of cockles in 2026 again now with cohorts
cockledata$FactCockleSize |>
  dplyr::filter(Year == 2026, !is.na(TransectPoint_ID),Present==1) |>
  ggplot2::ggplot(aes(x = Length_mm, fill=cohort)) +
  geom_histogram(binwidth = 1, boundary=0) +
  labs(x = "length of cockles (mm)", y = "number of cockles")


#Make a tibble with all combinations of transectpoints and replicate plots assessed for cockles in 2026, so that we can also include the transectpoints where no cockles were found in 2026.
design2026<-expand.grid(ReplicatePlot_ID=unique(cockledata$FactCockleSize$ReplicatePlot_ID[cockledata$FactCockleSize$Year == 2026]),
                        TransectPoint_ID=unique(cockledata$FactCockleSize$TransectPoint_ID[cockledata$FactCockleSize$Year == 2026])) |>
  dplyr::as_tibble()
design2026


# sum up  the number of cockles per transectpoint per replicate plot in 2026.
CockleCount2026present <- cockledata$FactCockleSize |>
  dplyr::filter(Year == 2026, !is.na(TransectPoint_ID),Present==1) |>
    dplyr::summarise(.by = c("TransectPoint_ID", "ReplicatePlot_ID", "cohort"), cockle_count = dplyr::n())
CockleCount2026present

# merge the design with the observation data, so that we also have the transect points where no cockles were found in 2026, and replace all `NA` in `n_obs` with zero, and also merge with the elevation data for 2026
CockleCount2026 <- dplyr::left_join(design2026, CockleCount2026present, by = c("TransectPoint_ID", "ReplicatePlot_ID")) |>
    dplyr::left_join(FactElevation2026, by = "TransectPoint_ID") |>
  dplyr::mutate(cockle_count = tidyr::replace_na(cockle_count, 0)) # replace na by 0

print(CockleCount2026)
```

Plot with a linear model how the number of cockles per transectpoint changes with elevation, all fonts size 10

```{r}
CockleCount2026 |> 
  dplyr::filter(Elevation_m > 0.25) |>  #remove outliers with very low elevation, not representative for the rest of the transect 
  ggplot2::ggplot(aes(x = Elevation_m, y = cockle_count)) +
  geom_point(aes(color = factor(ReplicatePlot_ID)), size = 1.5, alpha = 0.5) +
  geom_smooth(method = "lm", formula = "y~x") +
  labs(
    x = "Elevation (m + NAP)",
    y = "number of cockles per 0.068 m2",
    color = "replicate plot"
  ) +
  theme(text = element_text(size = 10))
```

Note the large variation between the replicate plots

# 04 Model the relation between cockle count and elevation

Explore how the cockle density changes with elevation.

```{r}
# Fit a linear regression: 
linreg <- stats::lm(cockle_count ~ Elevation_m, data = CockleCount2026)
summary(linreg)

# to test if the residuals show  a significant deviation from a normal distribution: 
stats::shapiro.test(residuals(linreg))

# This suggests that the residuals are not normally distributed, so the assumptions of the linear model are not met. We therefore fit a count model with a negative-binomial type 2 distribution, which allows the variance to increase quadratically with the mean.

glm1<-CockleCount2026 |>
  dplyr::filter(Elevation_m > 0.25) |>
  glmmTMB::glmmTMB(
    cockle_count ~ Elevation_m,
    family = stats::poisson(link = "log"),
    data = _
  ) 
summary(glm1) 

# check the conditions and assumptions of the model
simres_glm1 <- DHARMa::simulateResiduals(fittedModel = glm1, n = 1000)
DHARMa::testUniformity(simres_glm1)
# this doesn't look good, the residuals are not uniformly distributed, and there is a clear pattern in the residuals, indicating that the model does not fit the data well. 
DHARMa::testDispersion(simres_glm1)
# doesn't look good either, disperson test show significant overdispersion (variance>mean)
DHARMa::testZeroInflation(simres_glm1)
# this is ok - no significant zero-inflation

# check if the model improves if we add a quadratic term (second-order polynomial) to the model, to test for a unimodal relation between cockle count and elevation.

glm2<-CockleCount2026 |> 
  dplyr::filter(Elevation_m > 0.25) |>  #remove the outlier transect+id with very low elevation, not representative 
  glmmTMB::glmmTMB(
    cockle_count ~ Elevation_m + I(Elevation_m^2), 
    family = "poisson",
    data = _) 
summary(glm2) 
AIC(glm1,glm2)

# yes the quadratic term is significant, so we conclude that there is a unimodal relation between cockle count and elevation. Also the AIC is much lower for the quadratic model, indicating a better fit. 
# but we again want to  check the assumptions of this model: uniformity (q-q plot) and dispersion of residuals (mean=variance)

simres_glm2 <- DHARMa::simulateResiduals(fittedModel = glm2, n = 1000)
DHARMa::testUniformity(simres_glm2)
# this still doesn't look good, the residuals are not uniformly distributed, and there is a clear pattern in the residuals, indicating that the model does not fit the data well. 
DHARMa::testDispersion(simres_glm2)
# doesn't still look good either, disperson test show significant overdispersion (variance>mean)
DHARMa::testZeroInflation(simres_glm3)
# this is ok - no significant zero-inflation


# because of the lack of uniformity and overdispersion, we  try to fit a negative binomial model instead of a Poisson model that can better deal with these issues, as variance>mean of residuals

CockleCount2026 |>
  dplyr::filter(Elevation_m > 0.25) |>
  glmmTMB::glmmTMB(
    cockle_count ~ Elevation_m + I(Elevation_m^2),
    family = glmmTMB::nbinom2(link = "log"),
    data = _
  ) -> glm3
summary(glm3)
# test the dispersion of the residuals of the negative binomial model, to see if the model fits the data better than the Poisson model.
simres_glm3 <- DHARMa::simulateResiduals(fittedModel = glm3, n = 1000)
DHARMa::testDispersion(simres_glm3)
DHARMa::testUniformity(simres_glm3)
DHARMa::testZeroInflation(simres_glm3)
AIC(glm2,glm3)

# now all important assumptions of the GLM are met, so we can use this model to predict the number of cockles per transectpoint as a function of elevation.

# in case you want to check if adding the zero-inflation to the model would improve it (although there was no sign of it)
CockleCount2026 |>
  dplyr::filter(Elevation_m > 0.25) |>
  glmmTMB::glmmTMB(
    cockle_count ~ Elevation_m + I(Elevation_m^2),
    ziformula = ~1,
    family = glmmTMB::nbinom2(link = "log"),
    data = _
  ) -> glm4
summary(glm4)
# test the dispersion of the residuals of the negative binomial model, to see if the model fits the data better than the Poisson model.
simres_glm4 <- DHARMa::simulateResiduals(fittedModel = glm4, n = 1000)
DHARMa::testDispersion(simres_glm4)
DHARMa::testUniformity(simres_glm4)
DHARMa::testZeroInflation(simres_glm4)
AIC(glm3,glm4)
# AIC of this model is higher, the extra complexity is not justified by better fit

# add a random intercept for Transectpoint_ID (now we shift to a generalized linear mixed model!)
# This model accounts for the fact that observations from the same transect point are correlated. The fixed effects estimate the average quadratic relationship between elevation and cockle count, while the random intercept allows each transect point to have its own baseline count.
glm5 <- CockleCount2026 |>
  dplyr::filter(Elevation_m > 0.25) |>
  glmmTMB::glmmTMB(
    cockle_count ~ Elevation_m + I(Elevation_m^2) + (1 | TransectPoint_ID),
    family = glmmTMB::nbinom2(link = "log"),
    data = _
  )
summary(glm5)
DHARMa::testDispersion(simres_glm4)
DHARMa::testUniformity(simres_glm4)
DHARMa::testZeroInflation(simres_glm4)
AIC(glm3,glm5)


# glmmTMB is not available within geom_smooth, so we will use the glm3 model to predict the number of cockles per transectpoint as a function of elevation, and plot the predicted values along with the observed values.

# make a "prediction grid" with 200 values of elevation between the minimum and maximum elevation in the data, and predict the number of cockles per transectpoint for each value of elevation in the prediction grid, using the glm3 model.
prediction_grid <- data.frame(
  Elevation_m = seq(
    min(CockleCount2026$Elevation_m, na.rm = TRUE),
    max(CockleCount2026$Elevation_m, na.rm = TRUE),
    length.out = 200
  ),
  TransectPoint_ID = unique(CockleCount2026$TransectPoint_ID)[1]
) |>
  dplyr::filter(Elevation_m > 0.25)

# put the four different fitted models together in a list
models <- list(
  glm1 = glm1,
  glm2 = glm2,
  glm3 = glm3,
  glm5 = glm5
)

# add the predicted values of each of the three models to the prediction grid
prediction_grid[paste0("predicted_", names(models))] <- lapply(
  models,
  predict,
  newdata = prediction_grid,
  type = "response"
)

predictions_long <- prediction_grid |>
  tidyr::pivot_longer(
    cols = dplyr::starts_with("predicted_"),
    names_to = "model",
    values_to = "predicted"
  )
predictions_long

# Use distinct categorical colors for transect points and fitted models.
color_levels <- c(
  sort(unique(as.character(CockleCount2026$TransectPoint_ID))),
  unique(predictions_long$model)
)
set.seed(5831)
color_values <- setNames(
  sample(grDevices::colors(distinct = TRUE), length(color_levels)),
  color_levels
)

# plot the observation with the three models 
CockleCount2026 |>
  dplyr::filter(Elevation_m > 0.25) |>
  ggplot2::ggplot(aes(x = Elevation_m, y = cockle_count)) +
  geom_jitter(
    aes(color = factor(TransectPoint_ID)),
    width = 0.007,
    height = 0.15,
    size = 1.5
  ) +
  geom_line(
    data = predictions_long,
    aes(y = predicted, color = model),
    linewidth = 1.2
  ) +
  labs(
    x = "Elevation (m + NAP)",
    y = "number of cockles per 0.068 m2",
    color = "Transect point / model",
    title = "Predicted cockle counts as a function of elevation (glm5 is the preferred model)"
  ) +
  scale_color_manual(values = color_values) +
  theme(text = element_text(size = 10))

```

we find that the number of cockles is highest in the middle of this part of the transect, between transectpoint 170 and 2000, and that the relation between cockle count and elevation is unimodal. The only problem that we have not addressesed is that the data are not independent, as the same transectpoint is measured in three replicate plots, that are assumed to have the same elevation. We will address this issue in a next script, by fitting a generalized linear mixed model (GLMM) with transectpoint as a random effect.

