---
title: "Generalized linear mixed models"
subtitle: "Random intercepts, random slopes, and nested sampling designs"
author: "Han Olff"
date: 2025-09-21
format: html
editor_options:
  chunk_output_type: console
execute:
  warning: false
  message: false
---

# Introduction

Mixed-effects models extend ordinary linear models by allowing some model parameters to vary among groups of observations. Such grouping is often a deliberate part of the experimental design to control for factors that you know will have an effect, you want to control for that effect, but you are not specifically interested in the variation among those groups. For example, whe you do an experiment, you may group experimental treatment in replicate blocks, to control for spatial variation in the environment. Or you may do repeated measurement on the same plot or individual animal over time. In both examples, observations are not independent, they are grouped by some factor, and this grouping needs to be accounted for in the analysis. Mixed-effects models can be used to account for the non-independence of observations within suchgroups, and they can also be used to estimate these group-level effects. They are often call random effects, as you typically select the levels (blocks, plots, individuals) at random from a larger population of possible levels. In contrast, fixed effects are those that you are specifically interested in estimating, such as the effect of a treatment, an environmental variable or changes over time.

This document uses two examples. The first is a simulated warming experiment in the tundra with plant biomass after one season of warming, measured in plots of different treatments nested within blocks. It illustrates a random-intercept model, a random-slope model, and a model that also estimates the intercept–slope covariance. The second uses the `Salamanders` dataset from **glmmTMB** to illustrate nested random effects and repeated sampling.

# Setup

```{r}
#| label: setup-project
source(here::here("scripts", "01-setup.R"))

# load libraries
library(glmmTMB)
library(patchwork)
```

# Simulated warming experiment

## Experimental design

Suppose a tundra warming experiment has three blocks and four warming levels: 0 (control), 2, 4, and 8 degrees Celsius above ambient. The tundra is quite heterogenous. We therefore first select at random 10 different locations where we repeat the experiment. Each block has four subplots, one for each warming treatment. We then assign the warming treatments randomly to the subplots within each block. The experiment is run for one arctic summer, after which plant biomass is measured in each subplot. So each of the 4 warming treatments is replicated 10 times, because there are 10 blocks, and each block has one subplot for each warming treatment. The data are therefore grouped by block, and the blocks are expected to have different baseline biomass values, for example because of variation in vegetation or depth of the permafrost. But we do not try to quantify the causes of block differences (e.g., soil type, vegetation cover, etc.), we just want to account for the block-level variation in the analysis. if we would try to capture the differences between blocks in an explanatory variable, we replace the random effect with a fixed effect, but then we would have to measure **all** the relevant variables that explain the differences between blocks. In practice, this is often not possible and we do not know which ones, and we therefore use a random effect to account for block-level variation. So we just accept that block are different 'for some reason' and we want to account for that in the analysis.

The simulated values assume that warming has a linear effect within each block. The blocks have different baseline biomass values, but the first two simulated models differ in whether the warming slope is shared across blocks or allowed to vary among them. Residual variation is generated with a standard deviation of 5 g/m². We assume that the residuals are normally distributed, so we use a gaussian distribution with an identity link function.

```{r}
#| label: simulate-warming-data
# set the seed for reproducibility
set.seed(5831) 

treatments <- c(0, 2, 4, 6)
treatment_effects <- list(
  Block1 = c(5, 10, 15, 20),
  Block2 = c(5, 8, 12, 15),
  Block3 = c(20, 25, 27, 30)
)
block_means <- c(Block1 = 5, Block2 = 10, Block3 = 15)
n <- 10

warming_data <- purrr::map_dfr(
  names(treatment_effects),
  function(block) {
    purrr::map_dfr(seq_along(treatments), function(i) {
      treatment <- treatments[i]
      # Calculate the mean for this block-treatment combination
      mean_value <- treatment * treatment_effects[[block]][i] + block_means[[block]]

      tibble(
        Block = block,
        Treatment = treatment,
        Biomass_gm2 = rnorm(n, mean = mean_value, sd = 5)
      )
    })
  }
)

warming_data
```

## Explore the simulated data

We explore the results, where we jitter the point to avoid plotting points on top of eachother, allowing to see the variation better. And we fit a linear model to the data to visualize the overall trend. This model without the block effect will be our null model for comparison. We fit this directly with glmmTMB instead of lm() to keep the model fitting (maximum likelihood) consistent with the later models that include random effects. So we can statistically compare if more complex models with the random effects are better than the null model. The null model is a fixed-effects model with only the warming treatment as a predictor, and it does not account for block-level variation.

We see a clear positive effect of warming. But we have not accounted yet for the block-level variation, which is what we will do in the next section.

```{r}
#| label: plot-simulated-data-uncoloured

m0 <- glmmTMB::glmmTMB(
  Biomass_gm2 ~ Treatment,
  family = gaussian(link = "identity"),
  data = warming_data
)
m0
summary(m0)


warming_data |>
  ggplot(aes(x = Treatment, y = Biomass_gm2)) +
  geom_jitter(width = 0.05) +
  geom_smooth(method = "lm", formula = "y ~ x", color = "blue") +
  labs(
    x = "Warming treatment (°C enhancement)",
    y = "Biomass (g/m²)",
    title = "Simulated warming experiment without block colours"
  )
```

When we color the points by block, we see that the blocks have different baseline biomass values (biomass in the control plots, as 0 degree warming), and also the effect of warming slope may be different across blocks. This is what we will model in the next section. So we have to account also for blocks.

```{r}
#| label: plot-simulated-data

warming_data |>
  ggplot(aes(x = Treatment, y = Biomass_gm2, color = Block)) +
  geom_jitter(width = 0.05) +
  labs(
    x = "Warming treatment (°C enhancement)",
    y = "Biomass (g/m²)",
    title = "Simulated warming experiment"
  )
```

# Random-intercept model

A random-intercept mixed model fits for each block its own baseline biomass (intercept) while estimating one common warming slope:

$$
\text{Biomass}_{ij} = \beta_0 + \beta_1\text{Treatment}_{ij} + b_{0j} + \epsilon_{ij}.
$$

Here, $b_{0j}$ is the block-specific deviation from the overall intercept. The treatment effect is fixed, meaning that the model estimates one average slope for all blocks.

We assume that the residuals are normally distributed, so we use a gaussian distribution with an identity link function. We compare with anova() to the null model without the block effect if this model is significantly better. We also compare the AIC of the two models, where a lower AIC indicates a better trade-off between fit and model complexity.

```{r}
#| label: fit-random-intercept-model

m1 <- glmmTMB::glmmTMB(
  Biomass_gm2 ~ Treatment + (1 | Block),
  family = gaussian(link = "identity"),
  data = warming_data
)

summary(m1)
anova(m0, m1)
AIC(m0, m1)

```

The anova test finds that adding random intercepts for blocks significantly improves the model fit, and the AIC is much lower for the random-intercept model than for the null model. So the extra complexity is justified. 

## Plot random-intercept predictions

Predictions are calculated for each block so that the plot shows the different fitted intercepts while retaining the common fitted slope.

```{r}
#| label: predict-random-intercept-model

prediction_data <- tidyr::expand_grid(
  Treatment = seq(min(warming_data$Treatment), max(warming_data$Treatment), length.out = 100),
  Block = unique(warming_data$Block)
)

prediction_data$predicted <- predict(
  m1,
  newdata = prediction_data
)

warming_data |>
  ggplot(aes(x = Treatment, y = Biomass_gm2)) +
  geom_jitter(aes(color = Block), width = 0.1) +
  geom_line(
    data = prediction_data,
    aes(y = predicted, color = Block),
    linewidth = 1
  ) +
  labs(
    x = "Warming treatment (°C enhancement)",
    y = "Biomass (g/m²)",
    title = "Random-intercept model: common slope, block-specific intercepts"
  )
```

# Random-slope model

A random-slope model allows the warming response to vary among blocks:

$$
\text{Biomass}_{ij} = \beta_0 + \beta_1\text{Treatment}_{ij} + b_{0j} + b_{1j}\text{Treatment}_{ij} + \epsilon_{ij}.
$$

The model below estimates a random intercept and a random treatment slope. The `0 + Treatment` notation in the second random-effects term prevents that term from adding a second random intercept.

```{r}
#| label: fit-random-slope-model

m2 <- glmmTMB::glmmTMB(
  Biomass_gm2 ~ Treatment + (1 | Block) + (0 + Treatment | Block),
  family = gaussian(link = "identity"),
  data = warming_data
)

summary(m2)
car::Anova(m2)
AIC(m1, m2)
```

A lower AIC indicates a better trade-off between fit and model complexity. Differences of only a few AIC units should not be treated as decisive evidence on their own.

```{r}
#| label: predict-random-slope-model

prediction_data$predicted_m2 <- predict(
  m2,
  newdata = prediction_data
)

warming_data |>
  ggplot(aes(x = Treatment, y = Value)) +
  geom_jitter(aes(color = Block), width = 0.1) +
  geom_line(
    data = prediction_data,
    aes(y = predicted_m2, color = Block),
    linewidth = 1
  ) +
  labs(
    x = "Warming treatment (°C enhancement)",
    y = "Biomass (g/m²)",
    title = "Random-slope model: block-specific intercepts and slopes"
  )
```

## Random intercept–slope covariance

A more complex model allows the random intercept and random slope to be correlated. This asks whether blocks with higher baseline biomass also tend to have stronger or weaker warming responses.

```{r}
#| label: fit-correlated-random-effects-model

m3 <- glmmTMB::glmmTMB(
  Value ~ Treatment + (1 + Treatment | Block),
  family = gaussian(link = "identity"),
  data = warming_data
)

summary(m3)
```

With only three blocks, the data provide very little information for estimating a random-effects covariance reliably. A convergence warning or singular fit would therefore not be surprising, and this model should not be preferred without adequate replication.

```{r}
#| label: compare-random-effects-plots

p_random_intercept <- warming_data |>
  ggplot(aes(x = Treatment, y = Value)) +
  geom_jitter(aes(color = Block), width = 0.1) +
  geom_line(
    data = prediction_data,
    aes(y = predicted, color = Block),
    linewidth = 1
  ) +
  labs(title = "Random intercepts")

p_random_slope <- warming_data |>
  ggplot(aes(x = Treatment, y = Value)) +
  geom_jitter(aes(color = Block), width = 0.1) +
  geom_line(
    data = prediction_data,
    aes(y = predicted_m2, color = Block),
    linewidth = 1
  ) +
  labs(title = "Random slopes")

p_random_intercept / p_random_slope
```

# Nested sampling example

## Data structure

The `Salamanders` data contain repeated samples from sites. The same site can therefore contribute multiple observations, and samples are nested within sites. The data also include whether a site was mined and a measure of vegetation cover.

```{r}
#| label: load-salamanders-data

data(Salamanders, package = "glmmTMB")

dat <- Salamanders |>
  filter(spp == "DM") |>
  mutate(sample = factor(sample)) |>
  as_tibble()

dat
```

The following tables help inspect how sites, mining status, and samples are represented in the design.

```{r}
#| label: inspect-nested-design

table(dat$site, dat$mined)
table(dat$site, dat$sample)
```

## Explore the response and predictors

```{r}
#| label: plot-salamander-data

dat |>
  ggplot(aes(x = mined, y = count, color = sample)) +
  geom_boxplot()

dat |>
  ggplot(aes(x = site, y = count)) +
  geom_boxplot(aes(fill = site))

dat |>
  ggplot(aes(x = cover, y = count, color = site)) +
  geom_point(aes(shape = mined)) +
  geom_smooth(method = "lm")
```

# Fixed-effects count models

## Cover only

The first Poisson model estimates the association between vegetation cover and salamander count without accounting for mining status or site-level dependence.

```{r}
#| label: fit-cover-model

m4 <- glmmTMB::glmmTMB(
  count ~ cover,
  family = poisson(link = "log"),
  data = dat
)

summary(m4)
```

## Add mining status

The second model adds mining status as a fixed effect.

```{r}
#| label: fit-cover-mining-model

m5 <- glmmTMB::glmmTMB(
  count ~ cover + mined,
  family = poisson(link = "log"),
  data = dat
)

summary(m5)
car::Anova(m5)
```

# Add site-level dependence

## Random intercept for site

Because multiple samples were collected from the same site, observations from one site may be more similar than observations from different sites. A random site intercept accounts for this dependence by allowing each site to have its own baseline count.

```{r}
#| label: fit-site-random-intercept

m6 <- glmmTMB::glmmTMB(
  count ~ cover + mined + (1 | site),
  family = poisson(link = "log"),
  data = dat
)

summary(m6)
car::Anova(m6)
AIC(m5, m6)
```

## Add the nested sample effect

If the sampling design contains repeated sample identifiers within each site, a nested random-effects structure can be written as:

``` r
(1 | site) + (1 | site:sample)
```

This is equivalent to the shorthand:

``` r
(1 | site/sample)
```

The expanded form makes the two sources of variation explicit: variation among sites and additional variation among samples within sites.

```{r}
#| label: fit-nested-random-effects-model

m7 <- glmmTMB::glmmTMB(
  count ~ cover + mined + (1 | site) + (1 | site:sample),
  family = poisson(link = "log"),
  data = dat
)

summary(m7)
car::Anova(m7)
AIC(m6, m7)
```

The final comparison asks whether adding the nested sample-level random effect improves the model enough to justify the extra complexity. The answer should be based on the model comparison, convergence, variance estimates, and the sampling design—not on AIC alone.