---
title: "Generalized linear mixed models"
subtitle: "Random intercepts, random slopes, and nested sampling designs"
author: "Han Olff"
date: 2025-09-21
format: html
editor_options:
  chunk_output_type: console
execute:
  warning: false
  message: false
---

# Introduction

Mixed-effects models extend ordinary linear models by allowing some model parameters to vary among groups of observations. Such grouping is often a deliberate part of the experimental design to control for factors that you know will have an effect, you want to control for that effect, but you are not specifically interested in the variation among those groups. For example, when you do an experiment, you may group experimental treatment in replicate blocks, to control for spatial variation in the environment. Or you may do repeated measurement on the same plot or individual animal over time. In both examples, observations are not independent, they are grouped by some factor, and this grouping needs to be accounted for in the analysis. Mixed-effects models can be used to account for the non-independence of observations within suchgroups, and they can also be used to estimate these group-level effects. They are often call random effects, as you typically select the levels (blocks, plots, individuals) at random from a larger population of possible levels. In contrast, fixed effects are those that you are specifically interested in estimating, such as the effect of a treatment, an environmental variable or changes over time.

This document uses a examples of a simulated warming experiment in the tundra with plant biomass after one season of warming, measured in plots of different treatments nested within blocks. It illustrates a random-intercept model, a random-slope model, and a model that also estimates the intercept–slope covariance. 

# Setup

```{r}
#| label: setup-project
source(here::here("scripts", "01-setup.R"))

# load libraries
library(glmmTMB)
library(patchwork)
```

# Simulated warming experiment

## Experimental design

Suppose a tundra warming experiment has three blocks and four warming levels: 0 (control), 2, 4, and 8 degrees Celsius above ambient. The tundra is quite heterogenous. We therefore first select at random three different locations where we repeat the experiment. Within each block contains 30 plots with three different warming treatments (with cables in the soil), with 10 replicate plots per treatment. The warming treatments are randomly assigned to the replicates within each block. The experiment is run for one arctic summer, after which plant biomass is measured in each subplot. So each of the 4 warming treatments is replicated 30 times, because there are three blocks, and each block has 10 replicate subplots for each warming treatment. The data are therefore grouped by block, and the blocks are expected to have different baseline biomass values, for example because of variation in vegetation or depth of the permafrost. But we do not try to quantify the causes of block differences (e.g., soil type, vegetation cover, etc.), we just want to account for the block-level variation in the analysis. if we would try to capture the differences between blocks in an explanatory variable, we replace the random effect with a fixed effect, but then we would have to measure **all** the relevant variables that explain the differences between blocks. In practice, this is often not possible and we do not know which ones, and we therefore use a random effect to account for block-level variation. So we just accept that block are different 'for some reason' and we want to account for that in the analysis.

The simulated values assume that warming has a linear effect within each block. The blocks have different baseline biomass values, but the first two simulated models differ in whether the warming slope is shared across blocks or allowed to vary among them. Residual variation is generated with a standard deviation of 5 g/m². We assume that the residuals are normally distributed, so we use a gaussian distribution with an identity link function.

```{r}
#| label: simulate-warming-data
# set the seed for reproducibility
set.seed(5831) 

treatments <- c(0, 2, 4, 6)
treatment_effects <- list(
  Block1 = c(5, 10, 15, 20),
  Block2 = c(5, 8, 12, 15),
  Block3 = c(20, 25, 27, 30)
)
block_means <- c(Block1 = 5, Block2 = 10, Block3 = 15)
n <- 10

warming_data <- purrr::map_dfr(
  names(treatment_effects),
  function(block) {
    purrr::map_dfr(seq_along(treatments), function(i) {
      treatment <- treatments[i]
      # Calculate the mean for this block-treatment combination
      mean_value <- treatment * treatment_effects[[block]][i] + block_means[[block]]

      tibble(
        Block = block,
        Treatment = treatment,
        Biomass_gm2 = rnorm(n, mean = mean_value, sd = 5)
      )
    })
  }
)

warming_data
```

## Explore the simulated data

We explore the results, where we jitter the point to avoid plotting points on top of each other, allowing to see the variation better. And we fit a linear model to the data to visualize the overall trend. This model without the block effect will be our null model for comparison. We fit this directly with glmmTMB instead of lm() to keep the model fitting (maximum likelihood) consistent with the later models that include random effects. So we can statistically compare if more complex models with the random effects are better than the null model. The null model is a fixed-effects model with only the warming treatment as a predictor, and it does not account for block-level variation.

We see a clear positive effect of warming. But we have not accounted yet for the block-level variation, which is what we will do in the next section.

```{r}
#| label: plot-simulated-data-uncoloured

m0 <- glmmTMB::glmmTMB(
  Biomass_gm2 ~ Treatment,
  family = gaussian(link = "identity"),
  data = warming_data
)
m0
summary(m0)


warming_data |>
  ggplot(aes(x = Treatment, y = Biomass_gm2)) +
  geom_jitter(width = 0.05) +
  geom_smooth(method = "lm", formula = "y ~ x", color = "blue") +
  labs(
    x = "Warming treatment (°C enhancement)",
    y = "Biomass (g/m²)",
    title = "Simulated warming experiment without block colours"
  )
```

When we color the points by block, we see that the blocks have different baseline biomass values (biomass in the control plots, as 0 degree warming), and also the effect of warming slope may be different across blocks. This is what we will model in the next section. So we have to account also for blocks.

```{r}
#| label: plot-simulated-data

warming_data |>
  ggplot(aes(x = Treatment, y = Biomass_gm2, color = Block)) +
  geom_jitter(width = 0.05) +
  labs(
    x = "Warming treatment (°C enhancement)",
    y = "Biomass (g/m²)",
    title = "Simulated warming experiment"
  )
```

# Random-intercept model

A random-intercept mixed model fits for each block its own baseline biomass (intercept) while estimating one common warming slope:

$$
\text{Biomass}_{ij} = \beta_0 + \beta_1\text{Treatment}_{ij} + b_{0j} + \epsilon_{ij}.
$$

Here, $b_{0j}$ is the block-specific deviation from the overall intercept. The treatment effect is fixed, meaning that the model estimates one average slope for all blocks.

We assume that the residuals are normally distributed, so we use a gaussian distribution with an identity link function. We compare with anova() to the null model without the block effect if this model is significantly better. We also compare the AIC of the two models, where a lower AIC indicates a better trade-off between fit and model complexity, but on a gradual scale: How to interpret the difference in AIC:

| $\Delta AIC$ | Interpretation                         |
|--------------|----------------------------------------|
| 0–2          | Little difference between the models   |
| 2–4          | Some support for the lower-AIC model   |
| 4–7          | Clear support for the lower-AIC model  |
| \>7          | Strong support for the lower-AIC model |

```{r}
#| label: fit-random-intercept-model

m1 <- glmmTMB::glmmTMB(
  Biomass_gm2 ~ Treatment + (1 | Block),
  family = gaussian(link = "identity"),
  data = warming_data
)

summary(m1)
anova(m0, m1)
AIC(m0, m1)
AIC(m0)-AIC(m1)

```

The anova test finds that adding random intercepts for blocks significantly improves the model fit, and the delta-AIC (101 lower) is strong support for the random-intercept model compared to the null model without the block effect. So the extra complexity seems justified.

## Plot the predictions of the random intercept model

Predictions are calculated for each block so that the plot shows the different fitted intercepts while retaining the common fitted slope.

```{r}
#| label: predict-random-intercept-model

# predict the fitted values for each block across the range of warming treatments
prediction_data <- tidyr::expand_grid(
  Treatment = seq(min(warming_data$Treatment), max(warming_data$Treatment), length.out = 100),
  Block = unique(warming_data$Block)
)

prediction_data$predicted <- predict(
  m1,
  newdata = prediction_data
)

# plot the data and the fitted lines for each block
warming_data |>
  ggplot(aes(x = Treatment, y = Biomass_gm2)) +
  geom_jitter(aes(color = Block), width = 0.1) +
  geom_line(
    data = prediction_data,
    aes(y = predicted, color = Block),
    linewidth = 1
  ) +
  labs(
    x = "Warming treatment (°C enhancement)",
    y = "Biomass (g/m²)",
    title = "Biomass_gm2 ~ Treatment + (1 | Block)"
  )
```

This plot shows the so-called `random intercept model`. This name is a bit confusing. You see that every block has its own intercept, but the fitted slope (treatment effect) is the same for all blocks. So the intercepts are not selected at random, they are the fitted intercepts for each level of the random effect (block).

# Random-slope model

A random-slope model allows the warming response to vary among blocks, by fitting a slope for each block in addition to the block-specific intercept. So again the slopes are not random in a 'random-slope model', they are the slopes for each level of the random factor.

$$
\text{Biomass}_{ij} = \beta_0 + \beta_1\text{Treatment}_{ij} + b_{0j} + b_{1j}\text{Treatment}_{ij} + \epsilon_{ij}.
$$

The model below estimates a random intercept and a random treatment slope, while fixing their correlation to zero. The 0 in this notation prevents the model from estimating a correlation between the random intercept and random slope. This is often a good idea when the number of levels of the random effect is small, as in this example with only three blocks, because this covariance is then difficult to measure. A slightly shorter notation for this same model m2 would be ` Treatment + (Treatment || Block)`. But we use the more explicit form here as it is more clear, we just add the term (0 + Treatment | Block) to the previous model. The 1 in (1|block)indicates that the model estimates a random intercept for each block, and the 0 in (0 + Treatment | Block)indicates that the model estimates a random slope for each block but does not estimate a correlation between the random intercept and random slope (they are assumed to be independent, their covariance is fixed to zero). We fit the model again, test it with anova, and compare it with AIC to the previous model. 

```{r}
#| label: fit-random-slope-model

m2 <- glmmTMB::glmmTMB(
  Biomass_gm2 ~ Treatment + (1 | Block) + (0 + Treatment | Block),
  family = gaussian(link = "identity"),
  data = warming_data
)

summary(m2)
anova(m1,m2)
AIC(m1, m2)
AIC(m1)-AIC(m2)


```

So the model is again significant and there is strong support that the extra complexity is justified by more explained variation (delta-AIC is 136). Plot the model. In such mixed models, always show the fitted model equation when you show the graph and fit, so that the reader can see what is actually fitted. The fitted model equation is shown in the title of the plot below.

```{r}
#| label: predict-random-slope-model

prediction_data$predicted_m2 <- predict(
  m2,
  newdata = prediction_data
)

warming_data |>
  ggplot(aes(x = Treatment, y = Biomass_gm2)) +
  geom_jitter(aes(color = Block), width = 0.1) +
  geom_line(
    data = prediction_data,
    aes(y = predicted_m2, color = Block),
    linewidth = 1
  ) +
  labs(
    x = "Warming treatment (°C enhancement)",
    y = "Biomass (g/m²)",
    title = "Biomass_gm2 ~ Treatment + (1 | Block) + (0 + Treatment | Block)"
  )
```

## add a parameter to the model for the covariance between the random intercept and random slope

An even more complex model allows the random intercept and random slope to be correlated. This asks whether blocks with higher baseline biomass also tend to have stronger or weaker warming responses (as seems to be the case here). With only few levels of the random factor (3 in this case) we not always have enough power to fit this effect. But we will try. 

```{r}
#| label: fit-correlated-random-effects-model

m3 <- glmmTMB::glmmTMB(
  Biomass_gm2 ~ Treatment + (1 + Treatment | Block),
  family = gaussian(link = "identity"),
  data = warming_data
)
summary(m3)
anova(m2,m3)
AIC(m2, m3)
AIC(m2)-AIC(m3)
```

With only three blocks, the data provide very little information for estimating a random-effects covariance reliably. A convergence warning or singular fit would therefore not be surprising, and this model should not be preferred without adequate replication. 
Our anova tests shows that the models with the covariance term is not significantly better, and also the AIC test does not show sufficient support for the more complex model. So there is not enough support for including the covariance term in the model. Our model m2 with random intercepts and random slopes but no covariance term is therefore the best model for this dataset.

