---
title: "Generalized linear mixed models"
subtitle: "Random intercepts, random slopes, and nested sampling designs"
author: "Han Olff"
date: 2025-09-21
format: html
editor_options:
  chunk_output_type: console
execute:
  warning: false
  message: false
---

# Setup

```{r}
#| label: setup-project
source(here::here("scripts", "01-setup.R"))

# load libraries
library(glmmTMB)
library(patchwork)
```

The example study contains repeated counts of salamanders in mountain stream. The authors sampled 23 stream sites, each repeatedly sampled four times. Some sites were affected by mountaintop-removal coal mining, so a central question is how salamander abundance differs between mined and unmined streams.

# Data structure

The `Salamanders` data contain repeated samples from sites. The same site can therefore contribute multiple observations, and samples are nested within sites. The data also include whether a site was mined and a measure of vegetation cover. The repeated sampling is done not to assess temporal trends, but just because a single sampling round always has a risk to miss particular species. Repeated samples give a more complete view of each site.

Important variables:

| Variable | Meaning                                                          |
|------------------------------------|------------------------------------|
| `count`  | Number of salamanders observed                                   |
| `spp`    | Salamander species or life-stage category                        |
| `site`   | Sampling site                                                    |
| `mined`  | Whether the site was affected by mountaintop-removal coal mining |
| `cover`  | Amount of cover objects in the stream, scaled                    |
| `sample` | Repeated sampling occasion                                       |
| `DOP`    | Days since precipitation, scaled                                 |
| `Wtemp`  | Water temperature, scaled                                        |
| `DOY`    | Day of year, scaled                                              |

```{r}
#| label: load-salamanders-data

data(Salamanders, package = "glmmTMB")

dat <- Salamanders |>
  filter(spp == "DM") |>
  mutate(sample = factor(sample)) |>
  as_tibble()

dat
```

The following tables help inspect how sites, mining status, and samples are represented in the design.

```{r}
#| label: inspect-nested-design

table(dat$site, dat$mined)
table(dat$site, dat$sample)
```

## Explore the response and predictors

```{r}
#| label: plot-salamander-data

dat |>
  ggplot(aes(x = mined, y = count, color = sample)) +
  geom_boxplot()

dat |>
  ggplot(aes(x = site, y = count)) +
  geom_boxplot(aes(fill = site))

dat |>
  ggplot(aes(x = cover, y = count, color = site)) +
  geom_point(aes(shape = mined)) +
  geom_smooth(method = "lm")
```

# Fixed-effects count models

## Cover only

The first Poisson model estimates the association between vegetation cover and salamander count without accounting for mining status or site-level dependence.

```{r}
#| label: fit-cover-model

m1 <- glmmTMB::glmmTMB(
  count ~ cover,
  family = poisson(link = "log"),
  data = dat
)

summary(m1)
```

## Add mining status

The second model adds mining status as a fixed effect.

```{r}
#| label: fit-cover-mining-model

m2 <- glmmTMB::glmmTMB(
  count ~ cover + mined,
  family = poisson(link = "log"),
  data = dat
)

summary(m2)
car::Anova(m2)
anova(m1,m2)
AIC(m1,m2)
AIC(m1)-AIC(m2)
```

# Add site-level dependence

## Random intercept for site

Because multiple samples were collected from the same site, observations from one site may be more similar than observations from different sites; the observations are grouped by site. The sites may be different for various reasons other than the predictors in the model. A random site intercept accounts for this dependence by allowing each site to have its own baseline count.

```{r}
#| label: fit-site-random-intercept

m3 <- glmmTMB::glmmTMB(
  count ~ cover + mined + (1 | site),
  family = poisson(link = "log"),
  data = dat
)

summary(m3)
car::Anova(m3)
AIC(m2, m3)
```

## Add the nested sample effect

If the sampling design contains repeated sample identifiers within each site, a nested random-effects structure can be written as:

``` r
(1 | site) + (1 | site:sample)
```

This is equivalent to the shorthand:

``` r
(1 | site/sample)
```

The expanded form makes the two sources of variation explicit: variation among sites and additional variation among samples within sites.

```{r}
#| label: fit-nested-random-effects-model

m4 <- glmmTMB::glmmTMB(
  count ~ cover + mined + (1 | site) + (1 | site:sample),
  family = poisson(link = "log"),
  data = dat
)

# check assumptions of the final model as we did with a glm poisson model
simres_m4 <- DHARMa::simulateResiduals(fittedModel = m4, n = 1000)
DHARMa::testUniformity(simres_m4)
DHARMa::testDispersion(simres_m4)
DHARMa::testZeroInflation(simres_m4)


summary(m4)
car::Anova(m4)
AIC(m3, m4)
AIC(m3)-AIC(m4)
```

The final comparison asks whether adding the nested sample-level random effect improves the model enough to justify the extra complexity. The answer on the best model should be based on the check of the family of the model, model comparison, convergence, variance estimates, checks of assumptions etc and the sampling design—not on AIC alone. Use the similar checks as we did in script 11 for the generalized linear models.

Some further notes on the model specification:

1.  if locations within a site that are sampled at each sampling round are not permanent and not tracked, use as above:\
    `count ~ cover + mined + (1 | site) + (1 | site:sample)`

    \`Here:

    - `(1 | site)` accounts for repeated measurements from the same site.

    - `(1 | site:sample)` accounts for extra variation specific to each site during each sampling occasion, caused partly by sampling different locations.

2.  if the same samplingstations within each site are sampled repeatedly, and each has a permanent identifier 'samplingstation', that becomes an extra grouping factor and you model it explicitly as:

```         
`count ~ cover + mined +  (1 | site) +  (1 | site:samplingstation) +  (1 | sample)`

The term `(1 | site:location)` captures repeated measurements at the same location within a site, capturing spatial heterogeneity within each site. Not that the term `(1 | site:sample)` does not appear in the model anymore, as  extra variation specific to each site during each sampling occasion, caused by sampling different locations is not relevant anymore as you are now sampling the same locations within a site repeatedly. The term `(1 | sample)` now captures variation across sampling rounds, e.g., due to weather on each sampling day.
```

3.  For count responses, always check carefull for overdispersion, so if the variance of the **conditional response distribution** is larger than the mean. If so, use a negative binomial distribution:

    For a count response, also consider a negative-binomial model if the conditional response distribution is overdispersed (variance\>mean):

    `glmmTMB::glmmTMB( count ~ cover + mined + (1 | site) + (1 | site:sample),                                     family = glmmTMB::nbinom2)`

4. if your different sampling rounds aim to capture a temporal trends, then time becomes a fixed effect and you can model it for example as:

count ~ cover + mined + time + (1 | site) + (1 | site:samplingstation)
where you may also consider a random slope for time, if you test if  the temporal trend differs among sites: 
count ~ cover + mined + time +  (1 + time | site) +   (1 | site:samplingstation)


4.  What is the **conditional response distribution** in a Poisson regression (or in any glm or glmm? In a generalized linear model as a Poisson regression, the conditional response distribution describes how the response variable $Y_i$ is distributed for a given set of predictor values.

We write:

$$Y_i \mid \mathbf{x}_i \sim \operatorname{Poisson}(\mu_i)$$

This means that once the predictors $\mathbf{x}_i$ are fixed, the observed count $Y_i$ is assumed to come from a Poisson distribution with mean $\mu_i$.

For a Poisson distribution, the expected mean and variance are the same, so for a particular set of predictors $\mathbf{x}_i$

$$E(Y_i \mid \mathbf{x}_i) = \mu_i$$

and

$$\operatorname{Var}(Y_i \mid \mathbf{x}_i) = \mu_i.$$

So the key Poisson assumption is:

Conditional on the predictors, the variance of the response equals its mean.

The regression model then determines how this conditional mean depends on the predictors. With the usual log link:

\beta\_0 + \beta*1 x*{1i} + \beta*2 x*{2i} + \cdots \$\$

or equivalently,

\exp( \beta\_0 + \beta*1 x*{1i} + \beta*2 x*{2i} + \cdots ). \$\$

For example, suppose we model the number of salamanders as a function of tree cover:

$$Y_i \mid \text{treecover}_i
\sim
\operatorname{Poisson}(\mu_i)$$

with

\beta\_0 + \beta\_1 \text{treecover}\_i. \$\$

If the model predicts $\mu_i = 4$ (on average 4 salamanders) for a particular tree cover intensity, this does not mean that every plot with that tree cover contains exactly 4 species.

Instead, counts among such plots are assumed to follow a Poisson distribution with mean 4 and variance 4:

$$Y_i \sim \operatorname{Poisson}(4).$$

Individual plots may therefore contain, for example, 2, 3, 4, 5, or 6 species, while the expected number is 4 (the most likely number).

An important distinction is that the overall distribution of $Y$ (the number of salamanders counted) does not have to look Poisson. The assumption concerns the distribution of $Y$ conditional on the predictors; each tree cover is expected (if the variable is significant) to have a different count for salamanders.

So different observations can have different predictor values, giving different expected values $\mu_i$ and therefore different Poisson distributions.