---
title: "Generalized linear mixed models"
subtitle: "Random intercepts, random slopes, and nested sampling designs"
author: "Han Olff"
date: 2025-09-21
format: html
editor_options:
  chunk_output_type: console
execute:
  warning: false
  message: false
---

# Setup

```{r}
#| label: setup-project
source(here::here("scripts", "01-setup.R"))

# load libraries
library(glmmTMB)
library(patchwork)
```

The example study contains repeated counts of salamanders in mountain stream. The authors sampled 23 stream sites, each repeatedly sampled four times. Some sites were affected by mountaintop-removal coal mining, so a central question is how salamander abundance differs between mined and unmined streams.

# Data structure

The `Salamanders` data contain repeated samples from sites. The same site can therefore contribute multiple observations, and samples are nested within sites. The data also include whether a site was mined and a measure of vegetation cover. The repeated sampling is done not to assess temporal trends, but just because a single sampling round always has a risk to miss particular species. Repeated samples give a more complete view of each site.

Important variables:

| Variable | Meaning                                                          |
|------------------------------------|------------------------------------|
| `count`  | Number of salamanders observed                                   |
| `spp`    | Salamander species or life-stage category                        |
| `site`   | Sampling site                                                    |
| `mined`  | Whether the site was affected by mountaintop-removal coal mining |
| `cover`  | Amount of cover objects in the stream, scaled                    |
| `sample` | Repeated sampling occasion                                       |
| `DOP`    | Days since precipitation, scaled                                 |
| `Wtemp`  | Water temperature, scaled                                        |
| `DOY`    | Day of year, scaled                                              |

```{r}
#| label: load-salamanders-data

data(Salamanders, package = "glmmTMB")

dat <- Salamanders |>
  filter(spp == "DM") |>
  mutate(sample = factor(sample)) |>
  as_tibble()

dat
```

The following tables help inspect how sites, mining status, and samples are represented in the design.

```{r}
#| label: inspect-nested-design

table(dat$site, dat$mined)
table(dat$site, dat$sample)
```

## Explore the response and predictors

```{r}
#| label: plot-salamander-data

dat |>
  ggplot(aes(x = mined, y = count, color = sample)) +
  geom_boxplot()

dat |>
  ggplot(aes(x = site, y = count)) +
  geom_boxplot(aes(fill = site))

dat |>
  ggplot(aes(x = cover, y = count, color = site)) +
  geom_point(aes(shape = mined)) +
  geom_smooth(method = "lm")
```

# Fixed-effects count models

## Cover only

The first Poisson model estimates the association between vegetation cover and salamander count without accounting for mining status or site-level dependence.

```{r}
#| label: fit-cover-model

m1 <- glmmTMB::glmmTMB(
  count ~ cover,
  family = poisson(link = "log"),
  data = dat
)

summary(m1)
```

## Add mining status

The second model adds mining status as a fixed effect.

```{r}
#| label: fit-cover-mining-model

m2 <- glmmTMB::glmmTMB(
  count ~ cover + mined,
  family = poisson(link = "log"),
  data = dat
)

summary(m2)
car::Anova(m2)
anova(m1,m2)
AIC(m1,m2)
AIC(m1)-AIC(m2)
```

# Add site-level dependence

## Random intercept for site

Because multiple samples were collected from the same site, observations from one site may be more similar than observations from different sites; the observations are grouped by site. The sites may be different for various reasons other than the predictors in the model. A random site intercept accounts for this dependence by allowing each site to have its own baseline count.

```{r}
#| label: fit-site-random-intercept

m3 <- glmmTMB::glmmTMB(
  count ~ cover + mined + (1 | site),
  family = poisson(link = "log"),
  data = dat
)

summary(m3)
car::Anova(m3)
AIC(m2, m3)
```

## Add the nested sample effect

If the sampling design contains repeated sample identifiers within each site, a nested random-effects structure can be written as:

``` r
(1 | site) + (1 | site:sample)
```

This is equivalent to the shorthand:

``` r
(1 | site/sample)
```

The expanded form makes the two sources of variation explicit: variation among sites and additional variation among samples within sites.

```{r}
#| label: fit-nested-random-effects-model

m4 <- glmmTMB::glmmTMB(
  count ~ cover + mined + (1 | site) + (1 | site:sample),
  family = poisson(link = "log"),
  data = dat
)

# check assumptions of the final model as we did with a glm poisson model
simres_m4 <- DHARMa::simulateResiduals(fittedModel = m4, n = 1000)
DHARMa::testUniformity(simres_m4)
DHARMa::testDispersion(simres_m4)
DHARMa::testZeroInflation(simres_m4)


summary(m4)
car::Anova(m4)
AIC(m3, m4)
AIC(m3)-AIC(m4)
```

The final comparison asks whether adding the nested sample-level random effect improves the model enough to justify the extra complexity. The answer on the best model should be based on the check of the family of the model, model comparison, convergence, variance estimates, checks of assumptions etc and the sampling design—not on AIC alone. Use the similar checks as we did in script 11 for the generalized linear models. 