---
title: "14 Multivariate analysis"
subtitle: "DCA, CCA, PCA, PCoA, NMDS, and community classification"
author: "Han Olff"
date: "2025-09-16"
editor_options:
  chunk_output_type: console
execute:
  output: false
  warning: false
  message: false
---

# 01 Set up the environment

```{r}
# run setup script
source(here::here("scripts", "01-setup.R"))

# authenticate google sheets
gsheets_auth()

# load libraries needed for this script
library(vegan) # for DCA, RDA
library(psych)
library(ape) # for PCoA
library(ggrepel) # for avoiding overlapping labels in ggplot
library(ggvegan) # for plotting ordinations in ggplot

```

# 02 Read and combine the datasets

We will work with the vegetation data of 2026. First read the vegetation data into the `vegdat2026` dataframe. For ordination with functions from the vegan library you need an separate environmental factors dataset and a species composition dataset with the same rownames, indicating the same sites / samples the species data (community composition) need to be in wide format, so we produced this with `tibble::column_to_rownames()`. But do not convert it to a tibble, because the rownames are important for the ordination analyses and they are lost upon converion to a tibble. The environmental dataset will be created in the next code block, and will need to have the same rownames as the vegetation dataset, so that the ordination analyses can link the two datasets. In both files, the rownames are the plot numbers (TransectPoint_ID) and the columns are the species names (for the vegetation dataset) or the environmental variables (for the environmental dataset).

```{r}

# 1) read and filter the vegetation data for 2025
# browseURL("https://docs.google.com/spreadsheets/d/1WIzkUvunLk1buybZ04Oko8wnF65EbuXjWEviWbeAwu4")
VegetationDB<-read_gsdb("https://docs.google.com/spreadsheets/d/1WIzkUvunLk1buybZ04Oko8wnF65EbuXjWEviWbeAwu4",
                        sheet="FactVegCov")
vegdat2026<-VegetationDB$FactVegCov |>
  # filter for only the only 2026 data and plots with vegetation
  filter(Year==2026 & TransectPoint_ID<=1050) |>     
  # select  only distance_m and the species names as variable to use
  dplyr::select(-c(Year,Bare,Litter,Mosses,SalicSpp,totalcover)) |>
  # convert distance_m to the row names of the tibble (required by vegan)
  tibble::column_to_rownames(var="TransectPoint_ID")  %>%  
  # %>% needed because of  . representing the input data in next line
  # remove species are not found in any of the plots in this year
  dplyr::select(which(colSums(.) != 0)) 
  # this needs to stay a datafram (because of the row numbers, do not make it a tibble)
vegdat2026
```

Now read the different environmental variables from the separate databases and combine them with into a single dataset `envdat2026`. This dataset will contain the environmental variables for the same plots as in `vegdat2026`. Not that we can only use environmental variables that are measured for all plots, otherwise we will have missing values (NAs) in the dataset. We will remove those plots with NAs from the dataset, as the ordination analyses linking community composition to environmental variables require complete cases for both datasets. If you do not want that, you can also remove specific sites from the environmental dataset (that have NAs for some variable) but then you will have to remove those sites from the vegetation dataset as well.

```{r}
# dataset 1: macrotransect elevation data for 2026
MacrotransectDB<-read_gsdb("https://docs.google.com/spreadsheets/d/1UmpZoNIjo5lXdcpGKuBHe7wFX1Z3AQioeHjzrFgYnxo",
                           sheet=c("FactElevation","FactDist2Gully"))
MacrotransectDB$FactElevation2026 <- MacrotransectDB$FactElevation |>
  dplyr::filter(Year==2026 & !is.na(TransectPoint_ID) & TransectPoint_ID<=1050) |>
  dplyr::select(TransectPoint_ID,Elevation_m) |>   # select  only distance_m and elevation 
  dplyr::mutate(Elevation_m=round(Elevation_m,2))
MacrotransectDB$FactElevation2026

MacrotransectDB$FactDist2Gully <- MacrotransectDB$FactDist2Gully |>
  dplyr::filter(TransectPoint_ID<=1050) |>
  dplyr::select(TransectPoint_ID,Dist2Gully_m)   
print(MacrotransectDB$FactDist2Gully,n=25)

# dataset 2: transgression probability (proportion of the time of  the growing season flooded) of different elevations (per cm)
# from 1 april - 30 aug
SchierTideDB<-read_gsdb("https://docs.google.com/spreadsheets/d/1DOzvscotzWXm5MmEFrZhvPY820weEZVcSfUYz_5Gyf4",
                        sheets="FactTransProb") 
SchierTideDB$FactTransProb2026<-SchierTideDB$FactTransProb |>
  dplyr::filter(Year==2026) |>
  dplyr::select(Elevation_m,TransProb)
SchierTideDB$FactTransProb2026

# read the soil  database
SoilDB<-read_gsdb("https://docs.google.com/spreadsheets/d/1oHYCBlLR47Pxov5iF8PS_71lIRJwxgUfAP_t8JumJ7o",
                  sheet=c("FactProfile","FactRedox", "FactMoistureSalinity"))

# dataset 3: clay thickness
SoilDB$FactProfile2026<-SoilDB$FactProfile |>
  dplyr::filter(Year==2026 & SoilType_ID %in% c("clay","clay-organic") & TransectPoint_ID <= 1050) |>
  dplyr::select(TransectPoint_ID, corrected_depth)  |>     
  group_by(TransectPoint_ID) |> 
  dplyr::summarize(Clay_cm=mean(corrected_depth,na.rm=T)) #calculate average clay layer thickness  for each pole
SoilDB$FactProfile2026

# dataset 4:  redox
SoilDB$FactRedox2026<-SoilDB$FactRedox |>
  dplyr::filter(Year==2026 & TransectPoint_ID <= 1050) |>
  dplyr::group_by(TransectPoint_ID,ProbeDepth) |>
  dplyr::summarize(redox_mV=mean(redox_calib_mV,na.rm=T))  |>
  tidyr::pivot_wider(id_cols=TransectPoint_ID,
                     names_from=ProbeDepth,
                     names_prefix = "Redox",
                     values_from = redox_mV)
SoilDB$FactRedox2026

# dataset 5+6: moisture and conductivity
SoilDB$FactMoistureSalinity2026<-SoilDB$FactMoistureSalinity |>
  dplyr::filter(Year==2026 & TransectPoint_ID <= 1050 ) |>
  dplyr::group_by(TransectPoint_ID) |>
  dplyr::summarize(perc_water=mean(perc_water,na.rm=T),
                  conductivity=mean(conductivity_mSm,na.rm=TRUE)
)  
SoilDB$FactMoistureSalinity2026


# Merge all separate datafiles red so far in this codeblock into a single dataframe EnvDat
# use a sequential join operation in a pipe
# set distance_m as the row names of the  tibble
# set the NAs for floodprop equal to zero
# set the NAs for clay thickness equal to zero
# round floodprob to 2 decimals

envdat2026 <- 
  MacrotransectDB$FactElevation2026 |>
  dplyr::left_join(MacrotransectDB$FactDist2Gully, by="TransectPoint_ID") |>
  dplyr::left_join(SchierTideDB$FactTransProb2026, by="Elevation_m")  |>
  dplyr::left_join(SoilDB$FactProfile2026, by="TransectPoint_ID") |>
  dplyr::left_join(SoilDB$FactRedox2026, by="TransectPoint_ID") |>
  dplyr::left_join(SoilDB$FactMoistureSalinity2026, by="TransectPoint_ID") |>
  dplyr::mutate(TransProb=ifelse(is.na(TransProb),0,TransProb),
                TransProb=round(TransProb,3),
                Clay_cm=ifelse(is.na(Clay_cm),0,Clay_cm)) |>
  # convert to dataframe with row names required by vegan
  tibble::column_to_rownames(var="TransectPoint_ID")  |>
  #only retain complete cases - important!!
  tidyr::drop_na()
print(envdat2026)

```

## Explore the correlations among the environmental factors in a panel pairs plot

A pairs panel plot is a good way to inspect the correlations among the environmental factors. The diagonal shows the distribution of each variable, the lower triangle shows scatterplots of each pair of variables, and the upper triangle shows the correlation coefficients. Note that the units of the different environmental variables are very different, so it is important to use a correlation coefficient that is not sensitive to differences in scale, such as Spearman's rank correlation coefficient. This plot also helps you to identify collinearity among the environmental variables, which can be a problem in multivariate analyses. If two variables are highly correlated, you may want to remove one of them from the analysis. Also it helps you detect clear outliers, which may represent data entry mistakes or extreme values that may influence the results of the analysis. In that case, you may want to remove those outliers from the analysis.

```{r}
# using the pearson (linear) rank-correlation coefficient
psych::pairs.panels(envdat2026,smooth=F,ci=T,ellipses=F,stars=T,method="pearson")

```

# Ordination: run a Principal Component Analysis (PCA) on the environmental data

We start with a PCA on the environmental data, to reduce the dimensionality of the environmental data and to identify the main environmental gradients our study design. The PCA is a linear ordination method, which assumes that the environmental factors respond linearly to the environmental gradients. It is important to use .scale=T as a parameter of the PCA. The PCA is then based on the correlation matrix of the environmental variables, which means that the variables are standardized to have a mean of zero and a standard deviation of one. This is important because environmental variables are typically measured in different units and have widely different ranges. The PCA will produce a set of orthogonal axes (principal components) that explain the maximum amount of variation in the data. The first principal component (PC1) explains the most variation, followed by PC2, PC3, etc. You can see these as main environmental gradients represented by multiple environmental variables. We then visualize the PCA results with a biplot, which shows the scores of the sites (plots) and the loadings (scores) of the environmentalvariables on the first two principal components. The loadings indicate how much each variable contributes to each principal component. The biplot can help you identify which environmental variables are most important in explaining the variation in the data, which ones are highly correlated and how the sites (plots) are distributed along the main environmental gradients.

```{r}

# do a principal component analysis (pca) 
# note that an RDA without explanatory variables is equivalent to a PCA
# we use this for consistency with the other ordination methods (CCA, RDA, NMDS) for which we also use the vegan package
# this leads to consistency in the type of output and the way we can visualize the results

# remove Redox10,  Elevation_m from the environmental dataset (based on the outcome of the pairs.panels plot above) to avoid collinearity problems in the subsequent analyses
envdat2026uncor <- envdat2026 |>
  dplyr::select(-c(Redox10,Elevation_m))


pca_env<-vegan::rda(envdat2026uncor,center=T,scale=T)
pca_env
summary(pca_env)


## Make a biplot of the PCA results, with site codes as labels.
# extract species and sites scores from the ordination result
sites <- as.data.frame(vegan::scores(pca_env, display="sites", choices = 1:2))
sites$site <- rownames(sites)
species <- as.data.frame(vegan::scores(pca_env, display="species", choices = 1:2))
species$species <- rownames(species)

## Make the plot
ggplot() +
  geom_segment(
    data = species,
    aes(x = 0, y = 0, xend = PC1, yend = PC2),
    arrow = grid::arrow(length = grid::unit(0.15, "cm")),
    color = "red"
  ) +
  geom_text_repel(
    data = sites, col="blue", size=4,
    aes(x = PC1, y = PC2, label = site)
  ) +
  geom_text_repel(
    data = species, col="red", size=4,
    aes(x = PC1, y = PC2, label = species)
  ) +
  labs(x = "DCA1", y = "DCA2") +
  coord_equal() +
  theme_classic(base_size = 12) +
  theme(aspect.ratio = 1)

```

The plot show that the first principal component (PC1) explains 64% of the variation in the environmental data, and the second principal component (PC2) explains 16% of the variation. The arrows represent the loadings of the environmental variables on the principal components. All variables are about equal in terms of their contribution to the principal components. The length and direction of the arrows indicate the strength and direction of the relationship between each variable and the principal components. Not that several variables are strongly correlated, we may want to remove some of them from the analysis, to avoid collinearity problems in the subsequent analyses. For example, Redox5 and Redox10 are highly correlated, as are Elevation_m and TransProb. We will remove Redox10 and Elevation from the subsequent analyses. (see codeblock prior to pca)

# Ordination: DCA

A PCA is typically used for environmental data, which are typically continuous variables with often more or less linear relations. For species composition data, which are typically count data or presence-absence data, a unimodal ordination method such as DCA (Detrended Correspondence Analysis) is more appropriate. Unimodal relations of species with environmental gradients are expected from niche theory. DCA is a non-linear ordination method that assumes that species have unimodal responses to environmental gradients. The DCA can help you identify patterns in species composition and how they relate to environmental gradients. It is an indirect gradient analysis, which means that it does not use environmental variables to constrain the ordination, but rather uses the species composition data to identify the main gradients in the data. The DCA produces a set of orthogonal axes that explain the maximum amount of variation in the species composition data. The first axis (DCA1) explains the most variation, followed by DCA2, DCA3, etc. You can see these as main environmental gradients represented by multiple species. We then visualize the DCA results with a biplot, which shows the scores of the sites (plots) and the scores of the species on the first two DCA axes. The scores indicate how much each site or species contributes to each DCA axis. The biplot can help you identify which sites or species are most important in explaining the variation in the data, which ones are highly correlated and how the sites (plots) are distributed along the main environmental gradients, as inferred from the species composition data.

```{r}
# Keep the vegetation and environmental data aligned after removing incomplete sites.
complete_sites <- intersect(rownames(vegdat2026), rownames(envdat2026uncor))
complete_sites <- complete_sites[complete.cases(envdat2026uncor[complete_sites, , drop = FALSE])]
vegdat2026sel <- vegdat2026[complete_sites, , drop = FALSE]
envdat2026sel <- envdat2026uncor[complete_sites, , drop = FALSE]

# how long are the gradients? Should I use linear methods (PCA, RDA)or unimodal method (DCA, CCA)
dca<-vegan::decorana(vegdat2026sel)
dca
```

A rule of thumb is that when this length is \<3, you can use a PCA or RDA. When the length is \>4, you should use a unimodal method such as DCA or CCA. In this case, the length of the first axis is about 8 standard deviations of species responses, which indicates that the species composition data have a strong unimodal response to environmental gradients. Therefore, we will continue with DCA for the ordination of the species composition data. Not that in a DCA plot we do not use arrows for the species, as their score does not represent a linear gradient. Instead, the score of a species now represents their ecological optimum. Species that are close to each other in the DCA plot co-occur under the same environmental conditions, and species that are far apart have different ecological optima. The same applies to the sites (plots). Sites that are close to each other in the DCA plot have similar species composition, and sites that are far apart have different species composition. The DCA plot can help you identify patterns in species composition and how they relate to environmental gradients, as inferred from the species composition data.

```{r}
### plot the dca ordination with site codes as labels using ggvegan::autoplot
# extract species and sites scores from the ordination result
sites <- as.data.frame(scores(dca, display="sites", choices = 1:2))
sites$site <- rownames(sites)
species <- as.data.frame(scores(dca, display="species", choices = 1:2))
species$species <- rownames(species)

## Make the plot
ggplot() +
  geom_text_repel(
    data = sites, col="blue", size=4,
    aes(x = DCA1, y = DCA2, label = site)
  ) +
  geom_text_repel(
    data = species, col="red", size=4,
    aes(x = DCA1, y = DCA2, label = species)
  ) +
  labs(x = "DCA1", y = "DCA2") +
  coord_equal() +
  theme_classic(base_size = 12) +
  theme(aspect.ratio = 1)

```

## Fit environmental factors to the dca ordination surface

```{r}
names(envdat2026sel)
ef_dca <- vegan::envfit(
  dca ~ Dist2Gully_m + TransProb + Redox5 + Clay_cm + perc_water + conductivity,
  data = envdat2026sel,
  na.rm = FALSE
)

# Add environmental vectors to the DCA ordination. This takes a lot of script when we do it with ggvegan::autoplot, so we use the ordiplot function from vegan.
# Use a square plotting region for equal x- and y-axis dimensions.
par(pty = "s")
# make the ordination biplot with site codes as labels and add the environmental vectors in red
ordiplot(dca, display = "species", cex = 0.8, type = "text", xlim = c(-3, 3), ylim = c(-3, 3))
plot(ef_dca, add = TRUE, col = "blue")
```

## Add contour surfaces to the dca ordination for the relevant abiotic variables

```{r}
names(envdat2026sel)
vegan::ordiplot(dca,display="sites",cex=0.7,type="text",xlim=c(-3,3),ylim=c(-3,3))
vegan::ordisurf(dca, envdat2026sel$Redox5, add = TRUE, col = "red")
vegan::ordisurf(dca, envdat2026sel$conductivity, add = TRUE, col = "blue")

# make the plot again with contours of abundance of Plantago maritima
vegan::ordiplot(dca,display="sites",cex=0.7,type="text",xlim=c(-3,3),ylim=c(-1,1))
vegan::ordisurf(dca,vegdat2026sel$PlantMar,add=T,col="darkgreen")

```

# Ordination: CCA

A CCA is a good method if and which environmental factors jointly explain the variation in species composition among the sites. The CCA is a constrained ordination method, which means that it uses environmental variables to constrain the ordination. This means that the possible site scores on the axis are a multiple regression of the environmental predictors. You can see these as main environmental gradients (but only of the factors you measured) affecting multiple species. A CCA is especially good for (multivariate) hypothesis testing, as it allows you to test the significance of the environmental variables in explaining the variation in species composition.

```{r}
names(envdat2026sel)
cca<-vegan::cca(vegdat2026sel~Dist2Gully_m + TransProb + Redox5 + Clay_cm + perc_water + conductivity,
                data=envdat2026sel)
summary(cca)

# Test the whole model
anova(cca,permutations=9999)

# Test axes
anova(cca,by="axis", permutations=999)

# test variance inflation (values below 5 typically acceptable, values above 10 indicate collinearity problems)
vegan::vif.cca(cca)

# Test terms (environmental variables)
anova(cca,by="term", permutations=999)

cca<-vegan::cca(vegdat2026sel~Dist2Gully_m + TransProb + Redox5 + Clay_cm + perc_water + conductivity,
                data=envdat2026sel)

# plot the cca
# extract species, sites and env factor scores from the ordination result
species <- data.frame(vegan::scores(cca, display="species", choices = 1:2))
species$species <- rownames(species)
sites <- data.frame(vegan::scores(cca, display="sites", choices = 1:2))
sites$site <- rownames(sites)
environment <- as.data.frame(vegan::scores(cca, display = "bp", choices = 1:2))
environment$variable <- rownames(environment)

# Scale arrows for display; directions are interpreted in CCA space.
arrow_scale <- 0.8 / max(abs(as.matrix(environment[, c("CCA1", "CCA2")])) )
environment <- environment |>
  dplyr::mutate(
    CCA1 = CCA1 * arrow_scale,
    CCA2 = CCA2 * arrow_scale
  )

# make the plot with ggplot2
ggplot() +
  geom_segment(
    data = environment,
    aes(x = 0, y = 0, xend = CCA1, yend = CCA2),
    arrow = grid::arrow(length = grid::unit(0.15, "cm")),
    color = "darkgreen"
  ) +
  geom_text_repel(
    data = environment,
    aes(x = CCA1, y = CCA2, label = variable),
    color = "darkgreen",
    size = 4
  ) +
  geom_text_repel(
    data = sites,
    aes(x = CCA1, y = CCA2, label = site),
    color = "blue",
    size = 4
  ) +
  geom_text_repel(
    data = species,
    aes(x = CCA1, y = CCA2, label = species),
    color = "red",
    size = 4
  ) +
  labs(x = "CCA1", y = "CCA2") +
  coord_equal() +
  theme_classic(base_size = 12)
```


# ordination: PCoA
## Do a principal coordinate analysis PCoA

The PCA, DCA, and CCA are all ordination methods thatassume that the species abundances have a linear or unimodal response to the environmental gradients. When we use such methods, we explore what are the main environmental gradients affecting the species composition. However, in many cases, we are not so much interested in underlying gradients, but more in the similarity of species compostion between (groups of) sites. For example when we do a field experiment or when we want to compare the species composition between different land management regimes. In that case, a PCoA is a good method to use. PCoA is based on a distance matrix, which quantifies the dissimilarity in species composition between each pair of plots. The PCoA then tries to represent these dissimilarities in a low-dimensional space (usually 2D or 3D) while preserving the rank order of the dissimilarities as much as possible. Different dissimilarity measures can be used, depending on the type of data and the research question. 

```{r}
# calculate the distance matrix using Bray-Curtis dissimilarity
dist_mat <- vegan::vegdist(vegdat2026sel, method = "bray")
# calculate the PCoA using this distance matrix
pcoa_res <- vegan::wcmdscale(dist_mat, k = 2, eig = TRUE, add = "lingoes")

# Extract site coordinates and variance explained for plotting
sites <- as.data.frame(pcoa_res$points)
names(sites) <- c("PCoA1", "PCoA2")
sites$site <- rownames(sites)
var_exp <- round(pcoa_res$eig / sum(pcoa_res$eig) * 100, 1)[1:2]
sites

# a PCoA is only 'fed' with a dissimilarity matrix of sites. So the ordination result does not contain species scores. 
#  species scores can be approximated as weighted averages of site scores.
species <- as.data.frame(
  vegan::wascores(
    sites[, c("PCoA1", "PCoA2")],
    vegdat2026sel
  )
)
sites
species
species$species <- rownames(species)

# Fit environmental factors to the PCoA site scores.
spfit <- vegan::envfit(
  sites[, c("PCoA1", "PCoA2")],
  envdat2026sel,
  permutations = 999
)
environmentalfactors <- as.data.frame(
  vegan::scores(spfit, display = "vectors")
)
environmentalfactors$environmentalfactor <- rownames(environmentalfactors)



# Make the PCoA biplot.
ggplot2::ggplot() +
  ggplot2::geom_segment(
    data = environmentalfactors,
    ggplot2::aes(x = 0, y = 0, xend = PCoA1, yend = PCoA2),
    arrow = grid::arrow(length = grid::unit(0.15, "cm")),
    color = "darkgreen"
  ) +
  ggrepel::geom_text_repel(
    data = environmentalfactors,
    ggplot2::aes(PCoA1, PCoA2, label = environmentalfactor),
    color = "darkgreen",
    size = 4
  ) +
  ggplot2::geom_point(
    data = sites,
    ggplot2::aes(PCoA1, PCoA2),
    color = "blue",
    size = 2
  ) +
  ggrepel::geom_text_repel(
    data = sites,
    ggplot2::aes(PCoA1, PCoA2, label = site),
    color = "blue",
    size = 4
  ) +
  ggplot2::geom_point(
    data = species,
    ggplot2::aes(PCoA1, PCoA2),
    shape = 3,
    color = "red"
  ) +
  ggrepel::geom_text_repel(
    data = species,
    ggplot2::aes(PCoA1, PCoA2, label = species),
    color = "red",
    size = 4
  ) +
  ggplot2::labs(
    x = paste0("PCoA1 (", var_exp[1], "%)"),
    y = paste0("PCoA2 (", var_exp[2], "%)")
  ) +
  ggplot2::coord_equal() +
  ggplot2::theme_classic(base_size = 12) +
  ggplot2::theme(aspect.ratio = 1)


```

# Ordination: NMDS
An NMDS (non-metric multidimensional scaling) can be seen as a 'non-parametric' PCoA. Similar to a PcOA it orders sites in an ordination space where more similar sites are put closer together. But in contrast with PCOA, it does not use the actual dissimilarity but just the rank-order differences.  It is typicallyless informative than a DCA; as it visually overestimates often the importance of the second axis, and it does not provide information on the ecological optima of the species, the importance of axes, and of gradient lengths. All ordination axes are scaled to the same length, which is not the case in DCA. Therefore, we will prefer DCA for the ordination of the species composition data, and NMDS only as a complementary method to visualize the dissimilarity in species composition between plots.
vegan::metaNMDS() transforms the species abundances by default by the double Wisconsin-square root transformation. The square-root transformation reduces the influence of very abundant species. The Wisconsin standardization then gives species more comparable weight and removes differences in overall abundance among sites. This gives species with very different maximum abundances more equal influence. 


```{r}
# non-metric multidimension scaling / indirect gradient analysis (only species composition)
nmds_veg<-vegan::metaMDS(vegdat2026sel,k=2,trace=F,trymax=1000,distance="bray", autotransform = TRUE)
nmds_veg
vegan::ordiplot(nmds_veg,type="t")
# and show the ordination with the most abundance species with priority
SpecTotCov<-colSums(vegdat2026)
vegan::ordiplot(nmds_veg,display="sites",cex=1,type="t")
vegan::orditorp(nmds_veg,dis="sp",priority = SpecTotCov,
                col="red",pcol = "red",pch="+",cex=1.1)

# extract the species and site score
# extract species, sites and env factor scores from the ordination result
species <- data.frame(vegan::scores(nmds_veg, display="species", choices = 1:2))
species$species <- rownames(species)
sites <- data.frame(vegan::scores(nmds_veg, display="sites", choices = 1:2))
sites$site <- rownames(sites)
sites

# Fit the environmental factors to the nmds scores
environfit <- vegan::envfit(
  sites[, c("NMDS1", "NMDS2")],
  envdat2026sel,
  permutations = 999
)
environmentalfactors <- as.data.frame(
  vegan::scores(environfit, display = "vectors")
)
environmentalfactors$environmentalfactor <- rownames(environmentalfactors)

# Make the NMDS biplot.
ggplot2::ggplot() +
    ggplot2::geom_segment(
    data = environmentalfactors,
    ggplot2::aes(x = 0, y = 0, xend = NMDS1, yend = NMDS2),
    arrow = grid::arrow(length = grid::unit(0.15, "cm")),
    color = "darkgreen"
  ) +
  ggrepel::geom_text_repel(
    data = environmentalfactors,
    ggplot2::aes(NMDS1, NMDS2, label = environmentalfactor),
    color = "darkgreen",
    size = 4
  ) +
  ggplot2::geom_point(
    data = sites,
    ggplot2::aes(NMDS1, NMDS2),
    color = "blue",
    size = 2
  ) +
  ggrepel::geom_text_repel(
    data = sites,
    ggplot2::aes(NMDS1, NMDS2, label = site),
    color = "blue",
    size = 4
  ) +
  ggplot2::geom_point(
    data = species,
    ggplot2::aes(NMDS1, NMDS2),
    shape = 3,
    color = "red"
  ) +
  ggrepel::geom_text_repel(
    data = species,
    ggplot2::aes(NMDS1, NMDS2, label = species),
    color = "red",
    size = 4
  ) +
  ggplot2::labs(
    x = "NMDS1",
    y = "NMDS2"
  ) +
  ggplot2::coord_equal() +
  ggplot2::theme_classic(base_size = 12) +
  ggplot2::theme(aspect.ratio = 1)


```


# Cluster analysis (classification) of communities
In cluster analysis, we  classify sites into groups with similar species composition. In vegetation ecology, such group of sites or plots that have similar species composition are called **vegetation types**. This is a form of classification, which is different from ordination, which is a form of gradient analysis (rank sites and species into multiple dimensions base on their similarity). In cluster analysis, we first use a dissimilarity matrix to quantify the differences in species composition between each pair of sites. We then use a clustering algorithm to group the sites into clusters based on their dissimilarity. The result is a dendrogram, which shows the hierarchical relationships among the sites and the clusters. It shows how at different levels of similarity multiple sites are grouped together and at which level they split into subtypes. We can then cut the dendrogram at a certain level of dissimilarity to define the clusters, which then represent different communities with similar species composition. We can then use indicator species analysis to identify which species are characteristic of each cluster of site (so for each vegetation type).
```{r}
# first calculate a dissimilarity matrix, in this case we choose the  Bray-Curtis dissimilarity
d<-vegan::vegdist(vegdat2026sel,method="bray")
# show the dissimilarity matrix (1= completely different, 0= exactly the same)
d

# now cluster the sites based on similarity in species composition using "average linkage" method as the sorting algorithm; the rule we  use to decide how far apart two clusters are. Other methods can are also possible. 
cavg<-hclust(d,method="average")

# show the dendrogram 
plot(cavg)
# cut it into 5 communities (clusters) and show the clusters in the dendrogram
rect.hclust(cavg,5)
# store the cluster membership of each site in a new variable
clust<-cutree(cavg,5)
clust<-factor(clust)
clust

# do indicator species analysis - which species characterize each cluster? 
library(indicspecies)
set.seed(123) # for reproducibility
indval <- indicspecies::multipatt(vegdat2026sel, clust, 
                    func = "IndVal.g", duleg = F, 
                    control = how(nperm = 999))
summary(indval,indivalcomp=TRUE)
```

## Add the clustering of plots to your DCA ordination

```{r}
vegan::ordiplot(dca,display="sites",cex=0.7,type="text",xlim=c(-5,5))
vegan::orditorp(dca,dis="sp", priority=SpecTotCov,
                col="red",pcol="red",pch="+",cex=0.8,xlim=c(-5,5))
vegan::ordihull(dca,clust,lty=1,col="blue",lwd=2)

# add the vegetation type to the environmental data
envdat2026_2<-envdat2026sel |>
  dplyr::mutate(vegtype=clust)
envdat2026_2 |> as_tibble()


# name the different communities that you identified
levels(envdat2026_2$vegtype)<-c("Dune","High saltmarsh", 
                           "Mid Saltmarsh", "Low saltmarsh", "Pioneer")


# show boxplot of elevation differences between vegetation types
names(envdat2026_2)
p1<-envdat2026_2 %>% ggplot(aes(x=vegtype,y=Dist2Gully_m)) +
  geom_boxplot(fill="brown") +
  xlab(NULL) 
p1

# show boxplot of flooding probability differences between vegetation types
p2<-envdat2026_2 %>% ggplot(aes(x=vegtype,y=TransProb)) +
  geom_boxplot(fill="blue") +
  xlab(NULL) 
p2

# show boxplot of clay thickness differences between vegetation types
p3<-envdat2026_2 %>% ggplot(aes(x=vegtype,y=Clay_cm)) +
  geom_boxplot(fill="green") +
  xlab(NULL) 
p3

# show boxplot of redox at 5 cm between vegetation types
p4<-envdat2026_2 %>% ggplot(aes(x=vegtype,y=Redox5)) +
  geom_boxplot(fill="orange") +
  xlab(NULL) 
p4

p5<-envdat2026_2 %>% ggplot(aes(x=vegtype,y=perc_water)) +
  geom_boxplot(fill="purple") +
  xlab(NULL) 
p5

p6<-envdat2026_2 %>% ggplot(aes(x=vegtype,y=conductivity)) +
  geom_boxplot(fill="grey") 
p6


library(patchwork)
p1/p2/p3/p4/p5/p6

```

