---
title: "14 Multivariate analysis"
subtitle: "DCA, CCA, PCA, PCoA, NMDS, and community classification"
author: "Han Olff"
date: "2025-09-16"
editor_options:
  chunk_output_type: console
execute:
  output: false
  warning: false
  message: false
---

# 01 Set up the environment

```{r}
# run setup script
source(here::here("scripts", "01-setup.R"))

# authenticate google sheets
gsheets_auth()

# load libraries needed for this script
library(vegan) # for DCA, RDA
library(psych)
library(ape) # for PCoA
library(ggrepel) # for avoiding overlapping labels in ggplot

```

# 02 Read and combine the datasets

We will work with the vegetation data of 2026

```{r}

# 1) read and filter the vegetation data for 2025
# browseURL("https://docs.google.com/spreadsheets/d/1WIzkUvunLk1buybZ04Oko8wnF65EbuXjWEviWbeAwu4")
VegetationDB<-read_gsdb("https://docs.google.com/spreadsheets/d/1WIzkUvunLk1buybZ04Oko8wnF65EbuXjWEviWbeAwu4",
                        sheet="FactVegCov")
vegdat2026<-VegetationDB$FactVegCov |>
  # filter for only the only 2026 data and plots with vegetation
  filter(Year==2026 & TransectPoint_ID<=1050) |>     
  # select  only distance_m and the species names as variable to use
  dplyr::select(-c(Year,Bare,Litter,Mosses,SalicSpp,totalcover)) |>
  # convert distance_m to the row names of the tibble (required by vegan)
  tibble::column_to_rownames(var="TransectPoint_ID")  %>%  
  # %>% needed because of  . representing the input data in next line
  # remove species are not found in any of the plots in this year
  dplyr::select(which(colSums(.) != 0)) 
  # this needs to stay a datafram (because of the row numbers, do not make it a tibble)
vegdat2026

# 2) read and filter the macrotransect elevation data for 2026
MacrotransectDB<-read_gsdb("https://docs.google.com/spreadsheets/d/1UmpZoNIjo5lXdcpGKuBHe7wFX1Z3AQioeHjzrFgYnxo",
                           sheet=c("FactElevation","FactDist2Gully"))
MacrotransectDB$FactElevation2026 <- MacrotransectDB$FactElevation |>
  dplyr::filter(Year==2026 & !is.na(TransectPoint_ID) & TransectPoint_ID<=1050) |>
  dplyr::select(TransectPoint_ID,Elevation_m) |>   # select  only distance_m and elevation 
  dplyr::mutate(Elevation_m=round(Elevation_m,2))
MacrotransectDB$FactElevation2026

MacrotransectDB$FactDist2Gully <- MacrotransectDB$FactDist2Gully |>
  dplyr::filter(TransectPoint_ID<=1050) |>
  dplyr::select(TransectPoint_ID,Dist2Gully_m)   
print(MacrotransectDB$FactDist2Gully,n=25)

# 3) read the transgression probability (proportion of the time of  the growing season flooded) of different elevations (per cm)
# from 1 april - 30 aug
SchierTideDB<-read_gsdb("https://docs.google.com/spreadsheets/d/1DOzvscotzWXm5MmEFrZhvPY820weEZVcSfUYz_5Gyf4",
                        sheets="FactTransProb") 
SchierTideDB$FactTransProb2026<-SchierTideDB$FactTransProb |>
  dplyr::filter(Year==2026) |>
  dplyr::select(Elevation_m,TransProb)
SchierTideDB$FactTransProb2026


# 4) read the clay thickness and redox potential from the soil  database
SoilDB<-read_gsdb("https://docs.google.com/spreadsheets/d/1oHYCBlLR47Pxov5iF8PS_71lIRJwxgUfAP_t8JumJ7o",
                  sheet=c("FactProfile","FactRedox", "FactMoistureSalinity"))
SoilDB$FactProfile2026<-SoilDB$FactProfile |>
  dplyr::filter(Year==2026 & SoilType_ID %in% c("clay","clay-organic") & TransectPoint_ID <= 1050) |>
  dplyr::select(TransectPoint_ID,corrected_depth) |>     
  group_by(TransectPoint_ID) |> 
  dplyr::summarize(Clay_cm=mean(corrected_depth,na.rm=T)) #calculate average clay layer thickness  for each pole
SoilDB$FactProfile2026

# read redox
SoilDB$FactRedox2026<-SoilDB$FactRedox |>
  dplyr::filter(Year==2026 & TransectPoint_ID <= 1050) |>
  dplyr::group_by(TransectPoint_ID,ProbeDepth) |>
  dplyr::summarize(redox_mV=mean(redox_calib_mV,na.rm=T))  |>
  tidyr::pivot_wider(id_cols=TransectPoint_ID,
                     names_from=ProbeDepth,
                     names_prefix = "Redox",
                     values_from = redox_mV)
SoilDB$FactRedox2026

# read moisture and sality
SoilDB$FactMoistureSalinity2026<-SoilDB$FactMoistureSalinity |>
  dplyr::filter(Year==2026 & TransectPoint_ID <= 1050 ) |>
  dplyr::group_by(TransectPoint_ID) |>
  dplyr::summarize(perc_water=mean(perc_water,na.rm=T),
                  conductivity=mean(conductivity_mSm,na.rm=TRUE)
)  
SoilDB$FactMoistureSalinity2026

```

## Merge all previous 2025 datafiles into a single file EnvDat

```{r}
# a sequential join operation in a pipe
# set distance_m as the row names of the  tibble
# set the NAs for floodprop equal to zero
# set the NAs for clay thickness equal to zero
# round floodprob to 2 decimals

envdat2026 <- 
  MacrotransectDB$FactElevation2026 |>
  dplyr::left_join(MacrotransectDB$FactDist2Gully, by="TransectPoint_ID") |>
  dplyr::left_join(SchierTideDB$FactTransProb2026, by="Elevation_m")  |>
  dplyr::left_join(SoilDB$FactProfile2026, by="TransectPoint_ID") |>
  dplyr::left_join(SoilDB$FactRedox2026, by="TransectPoint_ID") |>
  dplyr::left_join(SoilDB$FactMoistureSalinity2026, by="TransectPoint_ID") |>
  dplyr::mutate(TransProb=ifelse(is.na(TransProb),0,TransProb),
                TransProb=round(TransProb,3),
                Clay_cm=ifelse(is.na(Clay_cm),0,Clay_cm)) |>
  # convert to dataframe with row names required by vegan
  tibble::column_to_rownames(var="TransectPoint_ID")  |>
  #only retain complete cases
  tidyr::drop_na()
print(envdat2026)  

# so for ordination with functions from the vegan library you need an separate environmental factors dataset
# and a species composition dataset with the same rownames, indicating  the same sites / samples
# the species data (community composition) need to be in wide format, so we produced this
print(vegdat2026)

```

## Explore the correlations among the environmental factors in a panel pairs plot

```{r}
# using the spearman rank-correlation coefficient
psych::pairs.panels(envdat2026,smooth=F,ci=T,ellipses=F,stars=T,method="spearman")
# note that the units are very different! 

```

## Ordination: run a Principal Component Analysis (PCA) on the environmental data

```{r}
# .scale=T means: use correlations instead of covariances
# use .scale=T for datasets where the variables are measured in different use

# do a principal component analysis (pca) 
pca_env<-prcomp(envdat2026,center=T,scale=T)
pca_env
summary(pca_env)

# make a biplot of the PCA results 
biplot(pca_env,scale=0,cex=c(0.7,1),xlab="PC1",ylab="PC2")

# you can also do this in ggplot, but this is substantially more work (but allows more flexibility in the plotting)
# The PCs are reduced dimensions of the environmental data.
pca_scores <- as.data.frame(pca_env$x[, 1:2]) |>
  tibble::rownames_to_column("site")
pca_scores

pca_loadings <- as.data.frame(pca_env$rotation[, 1:2]) |>
  tibble::rownames_to_column("variable")
pca_loadings

# Scale loading arrows to the range of the sample scores.
arrow_scale <- min(
  diff(range(pca_scores$PC1)) / diff(range(pca_loadings$PC1)),
  diff(range(pca_scores$PC2)) / diff(range(pca_loadings$PC2))
) * 0.7

pca_loadings <- pca_loadings |>
  dplyr::mutate(
    PC1 = PC1 * arrow_scale,
    PC2 = PC2 * arrow_scale
  )

explained_variance <- summary(pca_env)$importance[2, 1:2] * 100

ggplot2::ggplot(pca_scores, ggplot2::aes(x = PC1, y = PC2)) +
  ggplot2::geom_text(ggplot2::aes(label = site)) +
  ggplot2::geom_segment(
    data = pca_loadings,
    ggplot2::aes(x = 0, y = 0, xend = PC1, yend = PC2),
    arrow = grid::arrow(length = grid::unit(0.2, "cm")),
    color = "red",
    inherit.aes = FALSE
  ) +
  ggrepel::geom_text_repel(
    data = pca_loadings,
    ggplot2::aes(x = PC1, y = PC2, label = variable),
    color = "red",
    inherit.aes = FALSE,
    min.segment.length = 0
  ) +
  ggplot2::labs(
    x = sprintf("PC1 (%.1f%%)", explained_variance[1]),
    y = sprintf("PC2 (%.1f%%)", explained_variance[2]),
    title = "PCA biplot of environmental variables"
  ) +
  ggplot2::coord_fixed() +
  ggplot2::theme_bw() +
  ggplot2::theme(aspect.ratio = 1)


```

## Ordination: calculate and plot a Non-metric Multidimensional Scaling (NMDS) ordination

```{r}
# filter the vegetation data for those plots that are retained in envdat2026 dataset
vegdat2026 <- vegdat2026[rownames(vegdat2026) %in% rownames(envdat2026), ]

# explore the distance (dissimilarity) in species composition between plots
vegdat2026
d1<-vegan::vegdist(vegdat2026,method="euclidean") # Euclidean dissimilarity
d1
d2<-vegan::vegdist(vegdat2026,method="bray") # Bray-Curtis dissimilarity
d2


# non-metric multidimension scaling / indirect gradient analysis (only species composition)
nmds_veg<-metaMDS(vegdat2026,k=2,trace=F,trymax=1000,distance="bray")
nmds_veg
vegan::ordiplot(nmds_veg,type="t")
# and show the ordination with the most abundance species with priority
SpecTotCov<-colSums(vegdat2026)
vegan::ordiplot(nmds_veg,display="sites",cex=1,type="t")
vegan::orditorp(nmds_veg,dis="sp",priority = SpecTotCov,
                col="red",pcol = "red",pch="+",cex=1.1)

```

## Ordination: compare to a DCA -\> decide what ordination we should do, linear or unimodal?

```{r}
# how long are the gradients? Should I use linear (PCA)or unimodal method (NMDS, DCA)
dca<-vegan::decorana(vegdat2026)
dca
# first axis is ~10 standard deviations of species responses
# result: length of first ordination axis is >8 standard deviations
# only when <1.5 you can use a PCA or RDA
# Plot site and species scores using the same DCA scaling.
vegan::ordiplot(dca,display="sites",cex=0.7,type="text",xlim=c(-3,3),ylim=c(-3,3))
vegan::orditorp(dca,dis="sp", priority=SpecTotCov,
                col="red",pcol="red",pch="+",cex=0.8,xlim=c(-3,3),ylim=c(-3,3))
```

## Fit the environmental factors to the dca ordination surface

```{r}
names(envdat2026)
ef_dca<-vegan::envfit(dca~Clay_cm+Elevation_m+Dist2Gully_m+TransProb+Redox5+Redox10+perc_water+conductivity,
                      data=envdat2026,na.rm=T)
# add the result to the ordination plot as vectors for each variable
vegan::ordiplot(dca,display="sites",cex=0.7,type="text",xlim=c(-3,3),ylim=c(-3,3))
plot(ef_dca,add=T)
```

## Add contour surfaces to the dca ordination for the relevant abiotic variables

```{r}
vegan::ordiplot(dca,display="sites",cex=0.7,type="text",xlim=c(-3,3),ylim=c(-3,3))
vegan::ordisurf(dca,envdat2026$Dist2Gully_m,add=T,col="blue")
vegan::ordisurf(dca,envdat2026$TransProb,add=T,col="red")

# make the plot again with contours of abundance of Plantago maritima
vegan::ordiplot(dca,display="sites",cex=0.7,type="text",xlim=c(-3,3),ylim=c(-1,1))
vegan::ordisurf(dca,vegdat2026$PlantMar,add=T,col="red")

```

## Compare an unconstrainted (DCA) and constrained (CCA) ordination

```{r}
# did you miss important environmental factors?
# show the results of the detrended correspondence analysis

dca
# the eigenvalues represent the variation explained by each axis
# now do an cca
names(envdat2026)

cca<-vegan::cca(vegdat2026~Elevation_m+TransProb+Redox5+Clay_cm+perc_water+Dist2Gully_m,
                data=envdat2026)
summary(cca)

# Test the whole model
anova(cca,permutations=9999)

# Test axes
anova(cca,by="axis", permutations=999)

# Test terms (environmental variables)
anova(cca,by="term", permutations=999)


```

## Do a principal coordinate analysis PCoA

```{r}
dist_mat <- vegdist(vegdat2026, method = "bray")
pcoa_res <- ape::pcoa(dist_mat)
# extract the coordinates
scores <- as.data.frame(pcoa_res$vectors[,1:2])
names(scores) <- c("PCoA1","PCoA2")
scores$Plot_ID <- rownames(scores)   # rownames represent plot numbers in this case
scores
# calculate variance explained
var_exp <- round(pcoa_res$values$Relative_eig[1:2] * 100, 1)
# plot the ordination
# species "scores" as weighted averages of site scores (not that this is an approximation, not as in DCA or CCA)
sp <- vegan::wascores(scores[, c("PCoA1","PCoA2")], vegdat2026)
sp <- as.data.frame(sp)
sp$Species <- rownames(sp)
ggplot() +
  geom_point(data = scores, aes(PCoA1, PCoA2), size = 2) +
  geom_text_repel(data = scores, aes(PCoA1, PCoA2, label = Plot_ID), size = 3) +
  geom_point(data = sp, aes(PCoA1, PCoA2), shape = 3) +
  ggrepel::geom_text_repel(data = sp, aes(PCoA1, PCoA2, label = Species), size = 3) +
  xlab(paste0("PCoA1 (", var_exp[1], "%)")) +
  ylab(paste0("PCoA2 (", var_exp[2], "%)")) +
  theme_classic()

# repeat with a different dissimilarity measure, e.g. Hellinger
# 1. Hellinger transform
vegdat_hel <- decostand(vegdat, method = "hellinger")
# 2. PCA (=unconstrained RDA), when data are hellinger transformed this is equivalent to a PCoA with euclidean distance on the original data 
rda_res <- vegan::rda(vegdat_hel)
# 3. Site scores
scores_hel <- as.data.frame(scores(rda_res, display = "sites")[,1:2])
names(scores_hel) <- c("PCA1","PCA2")
scores_hel$Plot_ID <- rownames(scores_hel)
scores_hel
# 4. Species scores
sp_hel <- as.data.frame(scores(rda_res, display = "species")[,1:2])
names(sp_hel) <- c("PCA1","PCA2")
sp_hel$Species <- rownames(sp_hel)
sp_hel
# 5. % variance explained
var_exp_hel <- round(summary(rda_res)$cont$importance[2,1:2] * 100, 1)
# 6. Plot
ggplot() +
  geom_point(data = scores_hel, aes(PCA1, PCA2), size = 2) +
  geom_text_repel(data = scores_hel,
                  aes(PCA1, PCA2, label = Plot_ID),
                  size = 3) +
  geom_point(data = sp_hel, aes(PCA1, PCA2), shape = 3) +
  geom_text_repel(data = sp_hel,
                  aes(PCA1, PCA2, label = Species),
                  size = 3) +
  xlab(paste0("Axis 1 (", var_exp_hel[1], "%)")) +
  ylab(paste0("Axis 2 (", var_exp_hel[2], "%)")) +
  theme_classic()

```

## Solar park data

```{r}
solardat<-read_csv("https://docs.google.com/spreadsheets/d/e/2PACX-1vR0lG6_OUW42za4evantsfnerkHtud3Yb4T7IeRELBrKf6f3a04xwMXysX2wdPuZmPOGlyG9pBrcoE0/pub?gid=965225171&single=true&output=csv")



```

## Cluster analysis (classification) of communities

```{r}
# first calculate a dissimilarity matrix, using Bray-Curtis dissimilarity
d<-vegan::vegdist(vegdat,method="bray")
d

# show the dissimilarity matrix (1= completely different, 0= exactly the same)

# now cluster the sites based on similarity in species composition 
# using average linkage as the sorting algorithm
cavg<-hclust(d,method="average")
plot(cavg)


# show the dendrogram and cut it in 5 communities
rect.hclust(cavg,5)
clust<-cutree(cavg,5)
clust<-factor(clust)
clust

# do indicator species analysis - which species characterize each cluster? 
library(indicspecies)
set.seed(123) # for reproducibility
indval <- multipatt(vegdat, clust, 
                    func = "IndVal.g", duleg = F, 
                    control = how(nperm = 999))
summary(indval,indivalcomp=TRUE)



```

## Add the clustering of plots to your DCA ordination

```{r}
vegan::ordiplot(dca,display="sites",cex=0.7,type="text",xlim=c(-5,5))
vegan::orditorp(dca,dis="sp", priority=SpecTotCov,
                col="red",pcol="red",pch="+",cex=0.8,xlim=c(-5,5))
vegan::ordihull(dca,clust,lty=1,col="blue",lwd=2)



#add the vegetation type to the environmental data
envdat2<-envdat |>
  dplyr::mutate(vegtype=clust)
envdat2 |> as_tibble()


# name the different communities that you identified
levels(envdat2$vegtype)<-c("Dune","High saltmarsh", 
                           "Mid Saltmarsh", "Low saltmarsh", "Pioneer")


# show boxplot of elevation differences between vegetation types
p1<-envdat2 %>% ggplot(aes(x=vegtype,y=Elevation_m)) +
  geom_boxplot(fill="brown") +
  xlab(NULL) +
  theme(axis.text.x=element_blank(),
        axis.ticks = element_blank())
p1

# show boxplot of flooding probability differences between vegetation types
p2<-envdat2 %>% ggplot(aes(x=vegtype,y=TransProb)) +
  geom_boxplot(fill="blue") +
  xlab(NULL) +
  theme(axis.text.x = element_blank(),
        axis.ticks.x = element_blank())
p2

# show boxplot of clay thickness differences between vegetation types
p3<-envdat2 %>% ggplot(aes(x=vegtype,y=Clay_cm)) +
  geom_boxplot(fill="green") +
  xlab(NULL) +
  theme(axis.text.x = element_blank(),
        axis.ticks.x = element_blank())
p3

# show boxplot of distance to gully  between vegetation types
p4<-envdat2 %>% ggplot(aes(x=vegtype,y=Dist2Gully_m)) +
  geom_boxplot(fill="orange") +
  xlab(NULL) +
  theme(axis.text.x = element_blank(),
        axis.ticks.x = element_blank())
p4

p5<-envdat2 %>% ggplot(aes(x=vegtype,y=Redox5)) +
  geom_boxplot(fill="purple") +
  xlab(NULL) +
  theme(axis.text.x = element_blank(),
        axis.ticks.x = element_blank())
p5

p6<-envdat2 %>% ggplot(aes(x=vegtype,y=Redox10)) +
  geom_boxplot(fill="grey") 
p6


library(patchwork)
p1/p2/p3/p4/p5/p6


```