---
title: "Linear multiple regression with an interaction term"
subtitle: "Modeling clay depth as a function of elevation and year"
editor_options: 
  chunk_output_type: console
execute:
  output: false
  warning: false
  message: false
---

# 01 Setup

Run the common setup script and synchronize the local project with GitHub.

```{r}
# pull the changes in the online repository to this local folder
# this is important when you collaborate on the project
# it updates all your scripts to the latest version, and ensures that you have the same versions of the scripts as your collaborators
system2("git", c("-C", here::here(), "pull"))

# Include the setup script with source(). This will run the code in the setup script, and load the packages and functions that are defined in this script and that we will lead later on in upcoming scripts that we use in this course.
source(here::here("scripts", "01-setup.R"))

# load required libraries for this script
library(lm.beta)

#authenticate with Google Drive to read the dataset from the Google Sheet
googledrive::drive_auth(email = "h.olff@rug.nl")
```

# 02 Read and explore the dataset
Read the dataset from Google Sheets as a list of tibbles and copy the FactVegClay tibble to a local variable. The dataset contains clay depth measurements from the microtransect in the salt marsh, along with elevation and year of measurement.
```{r}
Microtransect <- read_gsdb("https://docs.google.com/spreadsheets/d/1dJkH09imko9RgOkGzYQT74IeGY56QwiXjjBl7u0KcK0/")
FactVegClay <- Microtransect$FactVegClay
FactVegClay
```

Explore the relation between clay depth and elevation, coloring points by year , and a linear regression line showing how overall ClayDepth_cm changes over with elevation (irrespective of year).

```{r}
FactVegClay |>
  ggplot2::ggplot(aes(x = Elevation_m, y = ClayDepth_cm)) +
  geom_point(aes(color = factor(Year))) +
  geom_smooth(method = "lm", se = FALSE)
```

The plot suggests that clay depth decreases with increasing elevation, but also that the relationship between clay depth and elevation may change over time (i.e., the increase seems to be stronger in more recent years). To investigate this further, we will have to fit a linear model with an interaction term between elevation and year.

But first also explore the relation between clay depth and year, coloring points by elevation with a diverging blue-yellow-red scale. 

```{r}
FactVegClay |>
  ggplot2::ggplot(aes(x = Year, y = ClayDepth_cm)) +
  geom_point(aes(color = Elevation_m)) +
  geom_smooth(method = "lm", se = FALSE) +
  scale_color_steps2(low = "blue", mid = "yellow", high = "red", 
                     midpoint = mean(FactVegClay$Elevation_m), n.breaks = 8)
```

The plots suggests that that clay depth tends to increase over time, but the relationship is not consistent across different elevations, at higher elevations the increase seems more pronounced. Again, to investigate this further, we will have to fit a linear model with an interaction term between elevation and year.

# 03 Fit lm model with interaction term
We will fit a linear model with an interaction term between elevation and year, to investigate whether the relationship between clay depth and elevation changes over time.
The model is Elevation_m + Year + Elevation_m:Year, where Elevation_m:Year is the interaction term between elevation and year. The model will be fitted to the data using the lm() function, and the results will be summarized using the broom package.
Note that this model can also be written als ClayDepth_cm ~ Elevation_m * Year, where the * operator indicates that both main effects and the interaction term should be included in the model.
```{r}
FactVegClay |>
  mutate(Elevation_m = Elevation_m/100) |> # scale elevation to 100 m units to make the coefficients more interpretable
  lm(ClayDepth_cm ~ Elevation_m + Year + Elevation_m:Year, data = _) -> lm_fit 

# show the overall fit of the model (R-squared, F-statistic, etc.) 
broom::glance(lm_fit)
# show the parameter estimates (coefficients) of the modeland their significance
broom::tidy(lm_fit)
# check the variance inflation factors (VIFs) of the model to check for multicollinearity
car::vif(lm_fit)
```

This output show various things, and how to continue depends on our interest. 
Four main aspects are relevant here:
-1) The overal fit of the model: The R-squared value indicates how much of the variance in the response variable (Claydepth) is explained by the model, 56% in this case. A higher R-squared value indicates a better fit.
-2) The significance of the model: The F-statistic and its associated p-value indicate whether the model is statistically significant, meaning that at least one of the predictors is significantly related to the response variable. In this case, the p-value is very small (p < 0.001), indicating that the model is statistically significant.
-3) The significance of the individual predictors: The coefficients table shows the estimated effect of each predictor on the response variable, along with their standard errors, t-values, and p-values. In this case, all three predictors (Elevation_m, Year, and the interaction term) are statistically significant (p < 0.001), indicating that they are all significantly explain variation in clay depth.
-4) Using the model to make predictions: the fitted model equation can be used to make predictions about clay depth for given values of Elevation_m and Year. The fitted model equation in which values for Elevation_m and Year can be plugged in to predict ClayDepth_cm is:
$$
\text{ClayDepth\_cm} = 4821.64 - 5091.16 \times \text{Elevation\_m} - 2.40 \times \text{Year} + 2.53 \times \text{Elevation\_m} \times \text{Year}
$$
where `Elevation_m` and `Year` are the original (uncentered) variables.
For example, for a point at 1.5 m elevation in 2010, the predicted clay depth is:
$$
\text{ClayDepth\_cm} = 4821.64 - 5091.16 \times 1.5 - 2.40 \times 2010 + 2.53 \times 1.5 \times 2010 = 4821.64 - 7636.74 - 4824 + 7615.95 = 976.85 \text{ cm}
$$
-5) Assess the relative importance of the different predictors; does elevation have a greater impact on clay depth than year? And how relatively important is the interaction term compared to the main effects? At first glace we might think that elevation (coefficient 5091) has much more impact than year (coefficient 2.40), and that the interaction term (coefficient 2.53) has about the same impact as year. But these cannot be compared between variables because the coefficients of this model are dependent of the measurement scale of the variables; they are "scale-dependent". The coefficient of 5091 for elevation means that for every 1 meter increase in elevation, the clay depth decreases by 5091 cm, holding all other predictors constant. So if we would not have expressed elevation in meters, but in centimeters, the coefficient of that elevation in the model would become 100 times larger, but of course not its relative importance relative to year. To compare coefficients between variables, we should look calculate their standardized coefficients (beta weights) from the model. These represent the change in the response variable (claydepth) for a one-standard-deviation change in each predictor, holding all other predictors constant. 
We can obtain these standardized coefficients by centering the predictors (subtracting the mean from each value) before fitting the model, and then using the `lm.beta()` function to obtain the standardized coefficients. We do this in the next section.
While doing this, we address an additional problem here: the interaction term between elevation and year is highly correlated with the main effects, which can lead to multicollinearity and unstable estimates. This is shown by the high value of the variance inflation factor (VIF) of 622927 for the interaction term, while a VIF of 1-5 is considered still ok. To address this, we will center the predictors before fitting the model in the next section.


# 04 Fit an lm model with standardized (beta) estimates to assess their relative importance

We will first center  `Elevation_m` and `Year` before fitting, based on the high VIFs calculated. This means that the mean of all elevations is subtracted from each elevation value, and the mean of all years is subtracted from each year value. This centers the data around zero, making the interpretation of the intercept more meaningful and reducing multicollinearity between the main effects and the interaction term. With an interaction term, leaving either predictor uncentered creates severe collinearity between the main effect and the interaction (inflated/misleading coefficients). These original calculated coeffisients are statistically unstable, meaning that a leaving a few observations out of the dataset or including some more would strongly change the value of the coefficients. Centering the predictors reduces this collinearity and stabilizes the coefficients, making them more interpretable.

So we will fit a single model on the centered (but not SD-scaled) predictors, then compute standardized coefficients from that same fit: `lm.beta()` takes the fitted `lm` object and adds a `standardized.coefficients` element to it, rather than requiring a separate refit on rescaled data. So `lm_fit_beta` is not a different model from `lm_fit` above — it *is* the centered `lm` fit, just with standardized coefficients attached, and it still works anywhere an `lm` object would (e.g. when used in `predict()`).

```{r}
FactVegClay |>
  # center year and elevation
  mutate(
    Elevation_m_c = Elevation_m - mean(Elevation_m),
    Year_c = Year - mean(Year)
  ) |>
  # calculate the multiple regression model with interaction term
  lm(ClayDepth_cm ~ Elevation_m_c * Year_c, data = _) |>
  # calculate the 
  lm.beta::lm.beta() -> lm_fit_beta

# inspect the standardized coefficients
broom::tidy(lm_fit_beta)
# inspect the overall fit of the model (R-squared, F-statistic, etc.)
broom::glance(lm_fit_beta)
```

Interpretation: standardized coefficients are ~0.56 for elevation and ~0.50 for year, indicating comparably strong associations with clay depth (in SD units) when each predictor is considered at the mean of the other. The interaction is much smaller (~0.15), suggesting the elevation-clay depth relationship changes only modestly across years, though it is still statistically significant (see p-value above). Note that standardizing the variables is especially important for their parameter estimate and standard error of that estimate, not so much for their significance (p-value) or overall model fit (R-squared). The p-values and R-squared are the same for the centered and uncentered models, but the coefficients are more interpretable when standardized.

# 05 plot the fitted model 

A naive way of proceeding would be to plot the model in ggplot for elevation versus clay-depths, with geom_smooth(method = "lm") combined year as a color aesthetic in the aes(). Shows that the effect of elevation is different for different years, sugggesting the interaction that we found in the model. 

```{r}
FactVegClay |>
  # lm_fit_beta$fitted.values are the unstandardized fitted values from
  # the underlying lm fit
  ggplot2::ggplot(aes(x = Elevation_m, y = ClayDepth_cm, color = factor(Year))) +
  geom_point() +
  geom_smooth(method = "lm", se = FALSE) +
  labs(x = "Elevation (m)", y = "Clay depth (cm)", color = "Year") +
  theme_bw()
```

But note that we are not the  model here that we just fitted. Instead we calculate here  a separate lm fit for each year! It does not accomodate year as having a linear effect on claydepth, as we tested in our model. To plot the fitted values from our model, we can use geom_line() with the fitted values from the centered lm fit `lm_fit_beta$fitted.values`. This works again because of row-order correspondence, not because the model is somehow evaluating uncentered predictors. `lm_fit_beta$fitted.values` is a vector of the already-computed fitted values from fitting on the centered data — one number per row of the training data `FactVegClay`, in the same row order.

```{r}
# plot the model with the fitted values from the centered lm fit
FactVegClay |>
  ggplot2::ggplot(aes(x = Elevation_m, y = ClayDepth_cm, color = factor(Year))) +
  geom_point() +
  geom_line(aes(y = lm_fit_beta$fitted.values), linewidth = 1) +
  labs(x = "Elevation (m)", y = "Clay depth (cm)", color = "Year") +
  theme_bw()

# predict clay depth at 1.05 m and 1.2 m elevation in 2017  and year 2025
predict(
  lm_fit_beta,
  newdata = data.frame(
    Elevation_m_c = c(1.05 - mean(FactVegClay$Elevation_m), 1.2 - mean(FactVegClay$Elevation_m)),
    Year_c = c(2017 - mean(FactVegClay$Year), 2025 - mean(FactVegClay$Year)),
    row.names = c("1.05", "1.2")
  )
)
```

Our final interpretation of the analysis is that sedimentation rate (increase in clay depth over time) is higher at higher elevations than at lower elevations. At 1.05 m elevation, clay depth increased by  0.97 cm over 8 years, while at 1.2 m elevation, clay depth increased  8.30 cm over the same period. This suggests that at low elevation some factor is inhibiting the sedimentation. This may be the sparse vegetation due to permanent water logging, leading to anoxia and high sulfide concentrations, and possibly also removal of sediment with microbial mats that dry up and are washed away. 

# 06 Accounting for repeated measures on the same transect points

`FactVegClay` is a balanced panel: 71 `Point_ID`s, each measured in every one of 9 years (639 = 71 × 9 rows). Observations from the same point across years are not independent, so the `lm()` models above understate the uncertainty of the fixed-effect estimates. We fit a mixed-effects model with `glmmTMB`, adding a per-point random intercept and random slope for `Elevation_m_c` (since `Elevation_m` changes over time within a point, its effect may vary by point too).

We considered adding an AR(1) residual correlation structure across years within each point (`glmm_fit_ar1`, using `ar1(Year_f + 0 | Point_ID)` with `Year` coded as a factor `Year_f`) to capture genuine temporal autocorrelation, beyond the general non-independence a random intercept/slope already accounts for. That combined model produces a convergence warning (non-positive-definite Hessian) and the residual variance collapses to near zero — with only 71 points × 9 years, there isn't enough replication to identify both a per-point random-slope covariance structure and a separate AR(1) term at once. We therefore drop the AR(1) term and instead compare two random-slope structures, both fit with a Gaussian family and identity link: one letting the per-point random intercept and slope correlate (`(Elevation_m_c | Point_ID)`, `glmm_fit_corr`), and a simpler one that fixes that correlation to zero (`(Elevation_m_c || Point_ID)`, `glmm_fit`). Both converge cleanly on their own (no AR(1) term); a likelihood-ratio test (`anova(glmm_fit, glmm_fit_corr)`) shows the intercept-slope correlation does not significantly improve the fit, so we prefer the simpler, uncorrelated model on parsimony grounds and inspect it with `summary()`.

```{r}
library(glmmTMB)

FactVegClay_glmm <- FactVegClay |>
  mutate(
    Elevation_m_c = Elevation_m - mean(Elevation_m),
    Year_c = Year - mean(Year),
    Year_f = factor(Year)
  )

# Random slope + AR(1) residual correlation 
glmm_fit_ar1 <- glmmTMB(
  ClayDepth_cm ~ Elevation_m_c * Year_c +
    (Elevation_m_c || Point_ID) +
    ar1(Year_f + 0 | Point_ID),
  family = gaussian("identity"),
  data = FactVegClay_glmm
)
summary(glmm_fit_ar1)
# But this does not converge reliably
# (non-positive-definite Hessian, residual variance collapses to ~0) --
# too many random-effect parameters for 71 points x 9 years of data.

# correlated random intercept and slope
glmm_fit_corr <- glmmTMB(
  ClayDepth_cm ~ Elevation_m_c * Year_c + (Elevation_m_c | Point_ID),
  family = gaussian("identity"),
  data = FactVegClay_glmm
)
broom::tidy(lm_fit_beta)
broom::glance(lm_fit_beta)


# uncorrelated random intercept and slope
glmm_fit <- glmmTMB(
  ClayDepth_cm ~ Elevation_m_c * Year_c + (Elevation_m_c || Point_ID),
  family = gaussian("identity"),
  data = FactVegClay_glmm
)
broom::tidy(lm_fit_beta)
broom::glance(lm_fit_beta)


# LRT: does allowing the intercept-slope correlation improve the fit?
anova(glmm_fit, glmm_fit_corr)

summary(glmm_fit)
```

Conclusion: repeated measures on the 71 transect points across 9 years mean observations aren't independent, so a plain `lm()` understates uncertainty. Three `glmmTMB` mixed models were compared: a random-slope-plus-AR(1) model (`glmm_fit_ar1`) failed to converge reliably (non-positive-definite Hessian, residual variance collapsing toward zero) since the data can't identify both a per-point slope covariance structure and a separate AR(1) term simultaneously. Of the two remaining models — with and without a correlation between the per-point random intercept and slope — a likelihood-ratio test found the correlation didn't significantly improve fit, so the simpler uncorrelated model (`glmm_fit`) was preferred. Its fixed-effect estimates (elevation ≈ 24.6, year ≈ 0.48, interaction ≈ 2.5, all significant) closely match the earlier `lm_fit_beta` estimates, but the standard errors here properly account for the repeated-measures structure rather than treating all 639 rows as independent.

# 07 Logistic regression of Limonium vulgare occurrence

`Limonium.vulgare` is a presence/absence (0/1) variable, so we model it with a binomial (logistic) glmmTMB model against elevation.
