---
title: "Linear multiple regression with an interaction term"
subtitle: "Modeling clay depth as a function of elevation and year"
editor_options: 
  chunk_output_type: console
execute:
  output: false
  warning: false
  message: false
---

# 01 Setup

Run the common setup script and synchronize the local project with GitHub.

```{r}
# pull the changes in the online repository to this local folder
# this is important when you collaborate on the project
# it updates all your scripts to the latest version, and ensures that you have the same versions of the scripts as your collaborators
system2("git", c("-C", here::here(), "pull"))

# Include the setup script with source(). This will run the code in the setup script, and load the packages and functions that are defined in this script and that we will lead later on in upcoming scripts that we use in this course.
source(here::here("scripts", "01-setup.R"))

# load required libraries for this script
library(lm.beta)

#authenticate with Google Drive to read the dataset from the Google Sheet
googledrive::drive_auth(email = "h.olff@rug.nl")
```

# 02 Read and explore the dataset
Read the dataset from Google Sheets as a list of tibbles and copy the FactVegClay tibble to a local variable. The dataset contains clay depth measurements from the microtransect in the salt marsh, along with elevation and year of measurement.
```{r}
Microtransect <- read_gsdb("https://docs.google.com/spreadsheets/d/1dJkH09imko9RgOkGzYQT74IeGY56QwiXjjBl7u0KcK0/")
FactVegClay <- Microtransect$FactVegClay
FactVegClay
```

Explore the relation between clay depth and elevation, coloring points by year , and a linear regression line showing how overall ClayDepth_cm changes over with elevation (irrespective of year).

```{r}
FactVegClay |>
  ggplot2::ggplot(aes(x = Elevation_m, y = ClayDepth_cm)) +
  geom_point(aes(color = factor(Year))) +
  geom_smooth(method = "lm", se = FALSE)
```

The plot suggests that clay depth decreases with increasing elevation, but also that the relationship between clay depth and elevation may change over time (i.e., the increase seems to be stronger in more recent years). To investigate this further, we will have to fit a linear model with an interaction term between elevation and year.

But first also explore the relation between clay depth and year, coloring points by elevation with a diverging blue-yellow-red scale. 

```{r}
FactVegClay |>
  ggplot2::ggplot(aes(x = Year, y = ClayDepth_cm)) +
  geom_point(aes(color = Elevation_m)) +
  geom_smooth(method = "lm", se = FALSE) +
  scale_color_steps2(low = "blue", mid = "yellow", high = "red", 
                     midpoint = mean(FactVegClay$Elevation_m), n.breaks = 8)
```

The plots suggests that that clay depth tends to increase over time, but the relationship is not consistent across different elevations, at higher elevations the increase seems more pronounced. Again, to investigate this further, we will have to fit a linear model with an interaction term between elevation and year.

# 03 Fit lm model with interaction term
We will fit a linear model with an interaction term between elevation and year, to investigate whether the relationship between clay depth and elevation changes over time.
The model is Elevation_m + Year + Elevation_m:Year, where Elevation_m:Year is the interaction term between elevation and year. The model will be fitted to the data using the lm() function, and the results will be summarized using the broom package.
Note that this model can also be written als ClayDepth_cm ~ Elevation_m * Year, where the * operator indicates that both main effects and the interaction term should be included in the model.
```{r}
FactVegClay |>
  lm(ClayDepth_cm ~ Elevation_m + Year + Elevation_m:Year, data = _) -> lm_fit 

broom::tidy(lm_fit)
broom::glance(lm_fit)
car::vif(lm_fit)
```

Fitted model equation:

$$
\text{ClayDepth\_cm} = 4821.64 - 5091.16 \times \text{Elevation\_m} - 2.40 \times \text{Year} + 2.53 \times \text{Elevation\_m} \times \text{Year}
$$

where `Elevation_m` and `Year` are the original (uncentered) variables.

Note: `Elevation_m` and the interaction term are severely collinear here (variance inflation factors — VIFs — in the hundreds of thousands) because `Year` is not centered, so these individual coefficients are numerically unstable and the coefficients of the model are not reliable. For the centered version with stable, interpretable estimates, see the next section.

# 05 Fit lm model with standardized (beta) estimates

Center `Elevation_m` and `Year` before fitting. With an interaction term, leaving either predictor uncentered creates severe collinearity between the main effect and the interaction (inflated/misleading coefficients), so both predictors must be centered for stable, interpretable estimates.

Fit the unstandardized model and the standardized (beta) version in one pipe: `lm.beta()` adds `standardized.coefficients` on top of the lm object, so `lm_fit_beta` still works anywhere `lm_fit` (an "lm") would, e.g. `predict()`.

```{r}
FactVegClay |>
  mutate(
    Elevation_m_c = Elevation_m - mean(Elevation_m),
    Year_c = Year - mean(Year)
  ) |>
  lm(ClayDepth_cm ~ Elevation_m_c * Year_c, data = _) |>
  lm.beta::lm.beta() -> lm_fit_beta

broom::tidy(lm_fit_beta)
broom::glance(lm_fit_beta)
str(lm_fit_beta)
```

```{r}
FactVegClay |>
  ggplot2::ggplot(aes(x = Elevation_m, y = ClayDepth_cm, color = factor(Year))) +
  geom_point() +
  geom_line(aes(y = lm_fit_beta$fitted.values), linewidth = 1) +
  labs(x = "Elevation (m)", y = "Clay depth (cm)", color = "Year") +
  theme_bw()
```

Interpretation: standardized coefficients are ~0.56 for elevation and ~0.50 for year, indicating comparably strong associations with clay depth (in SD units) when each predictor is considered at the mean of the other. The interaction is much smaller (~0.15), suggesting the elevation-clay depth relationship changes only modestly across years, though it is still statistically significant (see p-value above).

Check the difference if we would have used `geom_smooth()` with `lm` — this calculates a new lm fit for each year, so the results differ from the interaction model.

```{r}
FactVegClay |>
  # lm_fit_beta$fitted.values are the unstandardized fitted values from
  # the underlying lm fit
  ggplot2::ggplot(aes(x = Elevation_m, y = ClayDepth_cm, color = factor(Year))) +
  geom_point() +
  geom_smooth(method = "lm", se = FALSE) +
  labs(x = "Elevation (m)", y = "Clay depth (cm)", color = "Year") +
  theme_bw()
```

# 06 Logistic regression of Limonium vulgare occurrence

`Limonium.vulgare` is a presence/absence (0/1) variable, so we model it with a binomial (logistic) glmmTMB model against elevation.
