---
title: "Generalized linear mixed models"
subtitle: "Random intercepts, random slopes, and nested sampling designs"
author: "Han Olff"
date: 2025-09-21
format: html
editor_options:
  chunk_output_type: console
execute:
  warning: false
  message: false
---


# Setup

```{r}
#| label: setup-project
source(here::here("scripts", "01-setup.R"))

# load libraries
library(glmmTMB)
library(patchwork)
```




# Data structure

The `Salamanders` data contain repeated samples from sites. The same site can therefore contribute multiple observations, and samples are nested within sites. The data also include whether a site was mined and a measure of vegetation cover.

```{r}
#| label: load-salamanders-data

data(Salamanders, package = "glmmTMB")

dat <- Salamanders |>
  filter(spp == "DM") |>
  mutate(sample = factor(sample)) |>
  as_tibble()

dat
```

The following tables help inspect how sites, mining status, and samples are represented in the design.

```{r}
#| label: inspect-nested-design

table(dat$site, dat$mined)
table(dat$site, dat$sample)
```

## Explore the response and predictors

```{r}
#| label: plot-salamander-data

dat |>
  ggplot(aes(x = mined, y = count, color = sample)) +
  geom_boxplot()

dat |>
  ggplot(aes(x = site, y = count)) +
  geom_boxplot(aes(fill = site))

dat |>
  ggplot(aes(x = cover, y = count, color = site)) +
  geom_point(aes(shape = mined)) +
  geom_smooth(method = "lm")
```

# Fixed-effects count models

## Cover only

The first Poisson model estimates the association between vegetation cover and salamander count without accounting for mining status or site-level dependence.

```{r}
#| label: fit-cover-model

m4 <- glmmTMB::glmmTMB(
  count ~ cover,
  family = poisson(link = "log"),
  data = dat
)

summary(m4)
```

## Add mining status

The second model adds mining status as a fixed effect.

```{r}
#| label: fit-cover-mining-model

m5 <- glmmTMB::glmmTMB(
  count ~ cover + mined,
  family = poisson(link = "log"),
  data = dat
)

summary(m5)
car::Anova(m5)
```

# Add site-level dependence

## Random intercept for site

Because multiple samples were collected from the same site, observations from one site may be more similar than observations from different sites. A random site intercept accounts for this dependence by allowing each site to have its own baseline count.

```{r}
#| label: fit-site-random-intercept

m6 <- glmmTMB::glmmTMB(
  count ~ cover + mined + (1 | site),
  family = poisson(link = "log"),
  data = dat
)

summary(m6)
car::Anova(m6)
AIC(m5, m6)
```

## Add the nested sample effect

If the sampling design contains repeated sample identifiers within each site, a nested random-effects structure can be written as:

``` r
(1 | site) + (1 | site:sample)
```

This is equivalent to the shorthand:

``` r
(1 | site/sample)
```

The expanded form makes the two sources of variation explicit: variation among sites and additional variation among samples within sites.

```{r}
#| label: fit-nested-random-effects-model

m7 <- glmmTMB::glmmTMB(
  count ~ cover + mined + (1 | site) + (1 | site:sample),
  family = poisson(link = "log"),
  data = dat
)

summary(m7)
car::Anova(m7)
AIC(m6, m7)
```

The final comparison asks whether adding the nested sample-level random effect improves the model enough to justify the extra complexity. The answer should be based on the model comparison, convergence, variance estimates, and the sampling design—not on AIC alone.